# Community composition

## Metagenomics

```{r bats_load_data_tax, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
```

### Taxonomy overview 

***Number of Archaea phyla***
```{r arch_phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

***Number of Bacteria phyla***
```{r arch, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

#### Phylum level
```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Eptesicus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

***Phylum relative abundances***
```{r taxonomy_phylum_summary1, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,Species) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary2, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    Total_mean = mean(relabun * 100, na.rm = T),
    Total_sd = sd(relabun * 100, na.rm = T),
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total_meta = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Eptesicus_meta = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo_meta = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus_meta = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(phylum,
                Total_meta,
                Eptesicus_meta,
                Hypsugo_meta,
                Pipistrellus_meta) %>%
  tt()
```

#### Family level

***Percentange of families in each group***
```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  group_by(sample,family, Species) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_family_summary_sp, warning=FALSE, comments="", message=FALSE}
family_summary %>%
  group_by(family) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Eptesicus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean, -Ha_mean) %>%
  dplyr::select(family, Eptesicus, Hypsugo, Pipistrellus) %>%
  tt()

```

#### Genus level

***Percetange of genera in each group***
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, genus, Species) %>%
  summarise(relabun = sum(count))

genus_summary %>%
  group_by(genus) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Eptesicus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean, -Ha_mean) %>%
  dplyr::select(genus, Eptesicus, Hypsugo, Pipistrellus) %>% 
  tt()
```


***Number of mags and distinct taxonomy***
```{r tax_overview, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

total_mags <- data.frame(
  Bat = character(),
  MAGs = numeric(),
  Phylum = numeric(),
  Family = numeric(),
  Genus = numeric()
)

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

phylum <- preabs_table %>%
  distinct(phylum)

family <- preabs_table %>%
  distinct(phylum, class, order, family)

genus <- preabs_table %>%
  distinct(phylum, class, order, family, genus)

total_mags <- rbind(
  total_mags,
  data.frame(
    Bat = "Total",
    MAGs = nrow(preabs_table),
    Phylum = nrow(phylum),
    Family = nrow(family),
    Genus = nrow(genus)
  )
)

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  phylum <- preabs_table %>%
    select({{bat}}, phylum) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum)
  
  family <- preabs_table %>%
    select({{bat}}, phylum, class, order, family) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family, genus)
  
  total_mags <- rbind(
    total_mags,
    data.frame(
      Bat = bat,
      MAGs = nrow(number),
      Phylum = nrow(phylum),
      Family = nrow(family),
      Genus = nrow(genus)
    )
  )
}
```

```{r no_annot, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

no_annotation <- data.frame(Bat = character(),
                            No_genus = numeric(),
                            No_species = numeric())

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

genus <- preabs_table %>%
    filter(genus=="")


species <- preabs_table %>%
  filter(species=="")

no_annotation <- rbind(no_annotation,
                       data.frame(
                         Bat = "Total",
                         No_genus = nrow(genus),
                         No_species = nrow(species)
                       ))

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
  filter(genus=="")
  
  species <- preabs_table %>%
    filter(!!sym(bat) >= 1) %>%
  filter(species=="")
  
  no_annotation <- rbind(no_annotation,
                         data.frame(
                           Bat = bat,
                           No_genus = nrow(genus),
                           No_species = nrow(species)
                         ))
}
```

**Total percentage of MAGs without genus-level annotation**
```{r gen_percet, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
  filter(genus=="") %>% 
  summarize(Mag_nogenera = n()) %>%
  pull()
nmags <- total_mags %>%
  filter(Bat == "Total") %>%
  select(MAGs) %>%
  pull()
perct <- nongenera * 100 / nmags
cat(perct)
```

***Percentage of MAGs without genus-level annotation by phylum***
```{r nongene_phylum_gut, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(Total_MAGs = n())

genome_metadata %>%
    filter(genus=="")%>%
  group_by(phylum) %>%
  summarize(MAGs_nogenus = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Percentage_nogenus = 100 * MAGs_nogenus / Total_MAGs) %>%
  tt()
```

**Number of bacterial species**
```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(species) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Total percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nonspecies <- genome_metadata %>%
  filter(species=="") %>%
  summarize(Mag_nospecies = n()) %>%
  pull()
perct <- nonspecies * 100 / nmags
cat(perct)
```

**MAGs without species-level annotation**
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(MAGs_total = n())
genome_metadata %>%
  filter(species=="") %>%
  group_by(phylum) %>%
  summarize(MAGs_nospecies = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(species_annotated = MAGs_total - MAGs_nospecies) %>%
  mutate(Percentage_nospecies = 100 * MAGs_nospecies / MAGs_total) %>%
  mutate(Percentage_species = 100 - 100 * MAGs_nospecies / MAGs_total) %>%
  tt()
```

### Summary table
```{r single_sp_meta, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

single_sp <- data.frame(Bat = character(), Single_species = numeric())

table_upset_analysis <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample")

unique_all <- table_upset_analysis %>%
  filter(rowSums(across(Eb:Pk)) == 1)

single_sp <- rbind(single_sp, data.frame(Bat = "Total", Single_species = nrow(unique_all)))
  
for (bat in bats) {
  unique <- table_upset_analysis %>%
    filter(rowSums(across(Eb:Pk)) == 1) %>%
    select(all_of(bat)) %>%
    filter(. > 0) %>%
    nrow()
  
  single_sp <- rbind(single_sp, data.frame(Bat = bat, Single_species = unique))
}
```

```{r single_ind_meta, comment="", message=FALSE, warning=FALSE}
single_ind <- data.frame(Bat = character(), Single_individual = numeric())

freq_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("asv")

singleton_filt <- freq_table %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(-asv))) %>% 
  filter(row_sum == 1) %>%
  column_to_rownames(var = "asv")

single_ind <- rbind(single_ind, data.frame(
  Bat = "Total",
  Single_individual = nrow(singleton_filt) 
))

for (bat in bats) {
  single_filt <- singleton_filt %>%
    select(bat) %>%
    filter(. == 1)
  
  single_ind <- rbind(single_ind, data.frame(
    Bat = bat,
    Single_individual = nrow(single_filt)
  ))
}

```


```{r}
summary_table <- total_mags %>%
  left_join(., no_annotation, by = "Bat") %>%
  left_join(., single_ind, by = "Bat") %>%
  left_join(., single_sp, by = "Bat") %>% 
  mutate(MAGs_noSingleSpecies=MAGs-Single_individual)
summary_table
```

```{r}
summary_table %>%
  select(-Phylum, -Family, -Genus) %>%
  rowwise() %>%
  mutate(Mag_perc=MAGs*100/135)%>%
  mutate(MAGs_noSingleSp_perc=MAGs_noSingleSpecies*100/MAGs)%>% 
  mutate(No_genus_perc = No_genus * 100 / MAGs) %>%
  mutate(No_species_perc = No_species * 100 / MAGs) %>%
  mutate(Single_individual_perc = Single_individual * 100 / MAGs) %>%
  mutate(Single_species_perc = Single_species * 100 / MAGs) %>%
  mutate(Single_individual_per_Single_species = Single_individual * 100 /
           Single_species) %>%
  select(1,8:14)
```

## Amplicon
### Taxonomy overview 
```{r div_load_data_ampli, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```

#### Archaea
```{r}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```
```{r}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() 
```

```{r}
archaea_taxa <- taxa_names(phyloseq_16S_stand)[tax_table(phyloseq_16S_stand)[, "domain"] == "Archaea"]
phyloseq_16S_archaea <- prune_taxa(archaea_taxa, phyloseq_16S_stand)
phyloseq_16S_archaea <- prune_taxa(taxa_sums(phyloseq_16S_archaea)>0, phyloseq_16S_archaea)

#phyloseq_16S_archaea_pa <- transform_sample_counts(phyloseq_16S_archaea, function(x) as.integer(x > 0))

# Extract OTU table as a matrix
otu_pa <- otu_table(phyloseq_16S_archaea)

# Convert to presence-absence (1 if present, 0 if absent)
otu_pa[otu_pa > 0] <- 1

# Assign back to the phyloseq object, preserving taxa orientation
otu_table(phyloseq_16S_archaea) <- otu_table(otu_pa, taxa_are_rows = taxa_are_rows(phyloseq_16S_archaea))
phyloseq_16S_archaea <- prune_taxa(taxa_sums(phyloseq_16S_archaea)>0, phyloseq_16S_archaea)

otu_table(phyloseq_16S_archaea) %>% 
  rowSums()%>% 
  as.data.frame()  %>% 
  rename(count=1) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  group_by(phylum, genome) %>% 
  summarise(total_count=sum(count)) %>% 
  filter(total_count>1)

otu_table(phyloseq_16S_archaea) %>% 
  rowSums()%>% 
  as.data.frame()  %>% 
  rename(count=1) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  group_by(phylum, genus) %>% 
  summarise(total_count=sum(count)) %>% 
  filter(total_count>1) %>% 
  arrange(-total_count)
  
sample_sums(phyloseq_16S_archaea) %>% 
  as.data.frame() %>% 
  rename(count=1) %>% 
  filter(count>0) %>% 
  arrange(-count)

#filtering
sample_sums(phyloseq_16S_archaea) %>% 
  as.data.frame() %>% 
  rename(count=1) %>% 
  filter(count>0) %>% 
  arrange(-count) %>% 
  summarise(mean=mean(count),sd=sd(count))

#without filtering
sample_sums(phyloseq_16S_archaea) %>% 
  as.data.frame() %>% 
  rename(count=1) %>% 
  arrange(-count) %>% 
  summarise(mean=mean(count),sd=sd(count))
```
```{r}
sample_sums(phyloseq_16S_archaea) %>% 
  as.data.frame() %>% 
  rename(count=1) %>% 
  filter(count>0) %>% 
  arrange(-count) %>%
  mutate(across(everything(), ~ as.integer(. > 0)))%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata, by="sample")%>% 
  group_by(Species) %>% 
  summarise(n_individuals=sum(count))
```

```{r}
sample_sums(phyloseq_16S_archaea) %>% 
  as.data.frame() %>% 
  rename(count=1) %>% 
  filter(count>0) %>% 
  arrange(-count) %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata, by="sample")%>% 
  group_by(Species) %>% 
  summarise(total_count=sum(count))
```

Archaeal ASVs present in more than 1 individual
```{r}
data.frame(phyloseq_16S_archaea@otu_table) %>% 
  rowSums() %>% 
  as.data.frame() %>% 
  filter(.>1)
```


#### Phylum level
```{r taxonomy_barplot_ampli, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Eptesicus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 6),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 2))
```


***Phylum relative abundances***
```{r taxonomy_phylum_summary_ampli1, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, Species) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_phylum_summary_ampli2, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    Total_mean = mean(relabun * 100, na.rm = T),
    Total_sd = sd(relabun * 100, na.rm = T),
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Eptesicus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(phylum, Total, Eptesicus, Hypsugo, Pipistrellus) %>% 
  tt()
```

#### Family level

***Percentage of families in each group***
```{r taxonomy_family_summary_ampli, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  group_by(sample, family, Species) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_family_summary_ampli1, warning=FALSE, comments="", message=FALSE}
family_summary %>%
  group_by(family) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Eptesicus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Pk_mean) %>%
  dplyr::select(family, Eptesicus, Hypsugo, Pipistrellus) %>%
  left_join(., genome_metadata[c(3, 6)] %>% unique(), by = join_by(family == family)) %>% 
  tt()
```
```{r taxonomy_jitterplot_family_ampli, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
  group_by(family) %>%
  summarise(mean = sum(relabun)) %>%
  arrange(-mean) %>%
  select(family) %>%
  pull()

family_summary %>%
  left_join(genome_metadata %>% select(family, phylum) %>% unique(),
            by = join_by(family == family)) %>%
  filter(family %in% family_arrange[1:20]) %>%
  mutate(family = factor(family, levels = rev(family_arrange[1:20]))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(
    x = relabun,
    y = family,
    group = family,
    color = phylum
  )) +
  scale_color_manual(values = phylum_colors[-8]) +
  geom_jitter(alpha = 0.5) +
  facet_grid(. ~ Species) +
  theme_minimal() +
  labs(y = "Family", x = "Relative abundance", color = "Phylum")
```

#### Genus level

***Percetage of genera in each group***
```{r taxonomy_genus_summary_ampli, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, genus, Species) %>%
  summarise(relabun = sum(count))

genus_summary %>%
  group_by(genus) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Eptesicus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Pk_mean) %>%
  dplyr::select(genus, Eptesicus, Hypsugo, Pipistrellus) %>%
  left_join(., genome_metadata[c(3, 7)] %>% unique(), by = join_by(genus == genus)) %>% 
  tt()
```

```{r taxonomy_jitterplot_genus_ampli, fig.height=14, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
genus_summary_sort <- genus_summary %>%
  group_by(genus) %>%
  summarise(mean = mean(relabun, na.rm = T),
            sd = sd(relabun, na.rm = T)) %>%
  arrange(-mean)

genus_summary %>%
  mutate(genus = factor(genus, levels = rev(genus_summary_sort %>% pull(genus)))) %>%
  filter(relabun > 0) %>%
  ggplot(aes(
    x = relabun,
    y = genus,
    group = genus,
    color = phylum
  )) +
  scale_color_manual(values = phylum_colors) +
  geom_jitter(alpha = 0.5) +
  facet_grid(. ~ Species) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6)) +
  labs(y = "Genera", x = "Relative abundance", color = "Phylum")
```

***Number of ASVs and distinct taxonomy***
```{r total_ASv_ampli, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

total_asvs <- data.frame(
  Bat = character(),
  MAGs = numeric(),
  Phylum = numeric(),
  Family = numeric(),
  Genus = numeric()
)

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

phylum <- preabs_table %>%
  distinct(phylum)

family <- preabs_table %>%
  distinct(phylum, class, order, family)

genus <- preabs_table %>%
  distinct(phylum, class, order, family, genus)

total_asvs <- rbind(
  total_asvs,
  data.frame(
    Bat = "Total",
    ASVs = nrow(preabs_table),
    Phylum = nrow(phylum),
    Family = nrow(family),
    Genus = nrow(genus)
  )
)

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  phylum <- preabs_table %>%
    select({{bat}}, phylum) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum)
  
  family <- preabs_table %>%
    select({{bat}}, phylum, class, order, family) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family, genus)
  
  total_asvs <- rbind(
    total_asvs,
    data.frame(
      Bat = bat,
      ASVs = nrow(number),
      Phylum = nrow(phylum),
      Family = nrow(family),
      Genus = nrow(genus)
    )
  )
}
```

```{r total_annot_ampli, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

no_annotation <- data.frame(Bat = character(),
                            No_genus = numeric(),
                            No_species = numeric())

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

genus <- preabs_table  %>%
  filter(is.na(genus))

species <- preabs_table  %>%
  filter(is.na(species))

no_annotation <- rbind(no_annotation,
                       data.frame(
                         Bat = "Total",
                         No_genus = nrow(genus),
                         No_species = nrow(species)
                       ))

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(genus))
  
  species <- preabs_table %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(species))
  
  no_annotation <- rbind(no_annotation,
                         data.frame(
                           Bat = bat,
                           No_genus = nrow(genus),
                           No_species = nrow(species)
                         ))
}
```

**Total percentage of ASVs without genus-level annotation**
```{r gen_percet_ampli, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
  filter(is.na(genus)) %>% 
  summarize(ASV_nogenera = n()) %>%
  pull()
nasvs <- total_asvs %>%
  filter(Bat == "Total") %>%
  select(ASVs) %>%
  pull()
perct <- nongenera * 100 / nasvs
cat(perct)
```

***Percentage of ASVs without genus-level annotation by phylum***
```{r nongene_phylum_gut_ampli, comment="", message=FALSE, warning=FALSE}
total_asv_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(Total_ASVs = n())
genome_metadata %>%
  filter(is.na(genus)) %>% 
  group_by(phylum) %>%
  summarize(ASVs_nogenus = n()) %>%
  left_join(total_asv_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Percentage_nogenus = 100 * ASVs_nogenus / Total_ASVs) %>%
  tt()
```


**Number of bacterial species**
```{r phyla_ampli, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(species) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Total percentage of MAGs without species-level annotation**
```{r sp_percet_ampli, comment="", message=FALSE, warning=FALSE}
nonspecies <- genome_metadata %>%
     filter(is.na(species)) %>%
  summarize(ASV_nospecies = n()) %>%
  pull()
perct <- nonspecies * 100 / nasvs
cat(perct)
```

**MAGs without species-level annotation**
```{r nonspe_phylum_ampli, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(ASVs_total = n())
genome_metadata %>%
    filter(is.na(species)) %>%
  group_by(phylum) %>%
  summarize(ASVs_nospecies = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Species_annotated = ASVs_total - ASVs_nospecies) %>%
  mutate(Percentage_nospecies = 100 * ASVs_nospecies / ASVs_total) %>%
  mutate(Percentage_species = 100 - 100 * ASVs_nospecies / ASVs_total) %>%
  tt()
```

### Summary table

```{r single_sp_ampli, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

single_sp <- data.frame(Bat = character(), Single_species = numeric())

table_upset_analysis <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample")

unique_all <- table_upset_analysis %>%
  filter(rowSums(across(Eb:Pk)) == 1)

single_sp <- rbind(single_sp, data.frame(Bat = "Total",
                                         Single_species = nrow(unique_all)))
  
for (bat in bats) {
  unique <- table_upset_analysis %>%
    filter(rowSums(across(Eb:Pk)) == 1) %>%
    column_to_rownames(., "sample") %>% 
    select(bat) %>%
    filter(. > 0) %>%
    nrow()
  single_sp <- rbind(single_sp, data.frame(Bat = bat, Single_species = unique))
}

```


```{r single_ind_ampli, comment="", message=FALSE, warning=FALSE}
single_ind <- data.frame(Bat = character(), Single_individual = numeric())

freq_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("asv")

singleton_filt <- freq_table %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(-asv))) %>%
  filter(row_sum == 1) %>%
  column_to_rownames(var = "asv")  %>%
  filter(row_sum == 1)

single_ind <- rbind(single_ind, data.frame(
  Bat = "Total",
  Single_individual = nrow(singleton_filt)
))
     
for (bat in bats) {
  singleton_filt <- freq_table %>%
    rowwise() %>%
    mutate(row_sum = sum(c_across(-asv))) %>%
    filter(row_sum == 1) %>%
    column_to_rownames(var = "asv")  %>%
    select(bat) %>%
    filter(. == 1)
  
  single_ind <- rbind(single_ind, data.frame(
    Bat = bat,
    Single_individual = nrow(singleton_filt) 
  ))
}
```


```{r merge_ampli, comment="", message=FALSE, warning=FALSE}
summary_table_ampli <- total_asvs %>%
  left_join(., no_annotation, by = "Bat") %>%
  left_join(., single_ind, by = "Bat") %>%
  left_join(., single_sp, by = "Bat") %>% 
  mutate(ASVs_noSingleSpecies=ASVs-Single_individual)
  
summary_table_ampli
```


```{r summary_table_ampli, comment="", message=FALSE, warning=FALSE}
summary_table_ampli %>% 
  select(-Phylum,-Family, -Genus) %>% 
  rowwise() %>% 
  mutate(ASV_perc=ASVs*100/3834)%>% 
  mutate(ASVs_noSingleSp_perc=ASVs_noSingleSpecies*100/ASVs)%>% 
  mutate(No_genus_perc=No_genus*100/ASVs)%>% 
  mutate(No_species_perc=No_species*100/ASVs) %>% 
  mutate(Single_individual_perc=Single_individual*100/ASVs)%>% 
  mutate(Single_species_perc=Single_species*100/ASVs) %>% 
  mutate(Single_individual_per_unique=Single_individual*100/Single_species) %>% 
  select(1,8:14)
```


Phylum level
```{r}
load("resources/amplicon/data_standard_filt.Rdata")
genome_counts_filt_amplicon <- genome_counts_filt
genome_metadata_amplicon <- genome_metadata
load("resources/metagenomics/data_filtered.Rdata")
```

```{r}
topMAG_standard <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum) %>% 
  arrange(desc(total_mag))%>%
  slice_head(n = 10)
```


```{r}
topASV_standard <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 10)

table_standard <- full_join(topMAG_standard,topASV_standard, by="phylum") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
  select(-total_asv, -total_mag) %>% 
  pivot_longer(cols = c(MAG, ASV),
               names_to = "source",
               values_to = "abundance")
```


```{r figure_bar_phylum, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
ggplot(table_standard, aes(x = reorder(phylum, -abundance), y = abundance, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Phylum", y = "Total abundance", fill = "Source") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18))
```
```{r figure_bubble_phylum, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
ggplot(table_standard, aes(x = source, y = phylum, size = abundance, color = phylum)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(
    name = "Abundance",
    range = c(0.01, 30),   # 0 ensures no bubble at abundance = 0
    breaks = seq(0, 35, by = 5),   # Controls legend labels (your 0–5–10–...–30)
    limits = c(0.01, NA)   # keep full range of values
  )  +
    scale_color_manual(values = phylum_colors) +
    coord_flip()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))+
  labs(x = "Source", y = "Phylum",
    title = "Metabarcoding with standard filtering")+
  guides(color = "none", size = guide_legend(
      override.aes = list(size = c(2, 4, 6, 8, 10,12)) # smaller bubbles in legend
    )
  )
```

Family level
```{r}
topFMAG_standard <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum, family) %>% 
  arrange(desc(total_mag)) %>%
  slice_head(n = 20)
```

```{r}
topFASV_standard <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum, family) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 20)
```

```{r figure_scat_phylum, fig.height=8, fig.width=12, fig.fullwidth=TRUE}
full_join(topFMAG_standard,topFASV_standard, by="family") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
    mutate(phylum_mag = coalesce(phylum.x, phylum.y))%>%
  ggplot(., aes(x = MAG, y = ASV, color = phylum_mag, label = family)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
      scale_color_manual(values = phylum_colors) +
  labs(
    x = "MAG abundance",
    y = "ASV abundance",
    color = "Phylum",
    title = "Metabarcoding with standard filtering"
  ) +
  theme_minimal()
```