# Community composition

## Metagenomics

```{r bats_load_data_tax, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
```

### Taxonomy overview 
**Number of MAGs**
```{r mag_number, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  nrow()%>% 
  cat()
```

**Number of Archaea phyla**
```{r arch_phyla_meta, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

**Number of Bacteria phyla**
```{r bact_meta, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

#### Phylum level
```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```
**Grouping low-abundance bacteria**
```{r}
p1 <- genome_counts_filt %>%
  mutate_at(vars(-genome), ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>%
group_by(sample, phylum) %>%
  mutate(total_abundance = sum(count)) %>%
  ungroup() %>%
  mutate(phylum = if_else(total_abundance < 0.01, "Other", phylum)) %>%
  group_by(sample, phylum, Species) %>%
  summarise(count = sum(count), .groups = "drop") %>%
   ggplot(aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = c(phylum_colors, "Other" = "grey50"),
                    drop = FALSE)+
  facet_nested(~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6, face = "bold"),
    axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
#ggsave("community_plot_grouped_metagenomics.pdf", plot = p1, width = 12, height = 6)
p1
```

**Phylum relative abundances**
```{r taxonomy_phylum_summary_metag, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,Species) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_mmetag2, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    Total_mean = mean(relabun * 100, na.rm = T),
    Total_sd = sd(relabun * 100, na.rm = T),
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(phylum,
                Total,
                Cnephaeus,
                Hypsugo,
                Pipistrellus) %>%
  tt()
```

**Number of different phyla in each bat species**
```{r}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,domain,phylum,Species) %>%
  summarise(relabun=sum(count))

phylum_summary %>% 
  filter(relabun > 0) %>% 
  group_by(Species,domain) %>% 
  summarise(n_phyla = n_distinct(phylum)) %>% 
  tt()
```

#### Family level

**Percentange of families in each group**
```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  group_by(sample,family, Species) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_family_summary_sp, warning=FALSE, comments="", message=FALSE}
family_summary %>%
  group_by(family) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean, -Ha_mean) %>%
  dplyr::select(family, Cnephaeus, Hypsugo, Pipistrellus) %>%
  tt()

```

#### Genus level

**Percetange of genera in each group**
```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, genus, Species) %>%
  summarise(relabun = sum(count))

genus_summary %>%
  group_by(genus) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean, -Ha_mean) %>%
  dplyr::select(genus, Cnephaeus, Hypsugo, Pipistrellus) %>% 
  tt()
```

### MAGs

**Number of mags and distinct taxonomy**
```{r tax_overview, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

total_mags <- data.frame(
  Bat = character(),
  MAGs = numeric(),
  Phylum = numeric(),
  Family = numeric(),
  Genus = numeric()
)

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

phylum <- preabs_table %>%
  distinct(phylum)

family <- preabs_table %>%
  distinct(phylum, class, order, family)

genus <- preabs_table %>%
  distinct(phylum, class, order, family, genus)

total_mags <- rbind(
  total_mags,
  data.frame(
    Bat = "Total",
    MAGs = nrow(preabs_table),
    Phylum = nrow(phylum),
    Family = nrow(family),
    Genus = nrow(genus)
  )
)

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  phylum <- preabs_table %>%
    select({{bat}}, phylum) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum)
  
  family <- preabs_table %>%
    select({{bat}}, phylum, class, order, family) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family, genus)
  
  total_mags <- rbind(
    total_mags,
    data.frame(
      Bat = bat,
      MAGs = nrow(number),
      Phylum = nrow(phylum),
      Family = nrow(family),
      Genus = nrow(genus)
    )
  )
}
```

```{r no_annot, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

no_annotation <- data.frame(Bat = character(),
                            No_genus = numeric(),
                            No_species = numeric())

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))

genus <- preabs_table %>%
    filter(genus=="")


species <- preabs_table %>%
  filter(species=="")

no_annotation <- rbind(no_annotation,
                       data.frame(
                         Bat = "Total",
                         No_genus = nrow(genus),
                         No_species = nrow(species)
                       ))

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
  filter(genus=="")
  
  species <- preabs_table %>%
    filter(!!sym(bat) >= 1) %>%
  filter(species=="")
  
  no_annotation <- rbind(no_annotation,
                         data.frame(
                           Bat = bat,
                           No_genus = nrow(genus),
                           No_species = nrow(species)
                         ))
}
```

**Total percentage of MAGs without genus-level annotation**
```{r gen_percet, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
  filter(genus=="") %>% 
  summarize(Mag_nogenera = n()) %>%
  pull()
nmags <- total_mags %>%
  filter(Bat == "Total") %>%
  select(MAGs) %>%
  pull()
perct <- nongenera * 100 / nmags
cat(perct)
```

**Percentage of MAGs without genus-level annotation by phylum**
```{r nongene_phylum_gut, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(Total_MAGs = n())

genome_metadata %>%
    filter(genus=="")%>%
  group_by(phylum) %>%
  summarize(MAGs_nogenus = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Percentage_nogenus = 100 * MAGs_nogenus / Total_MAGs) %>%
  tt()
```

**Number of bacterial species**
```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(species) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Total percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nonspecies <- genome_metadata %>%
  filter(species=="") %>%
  summarize(Mag_nospecies = n()) %>%
  pull()
perct <- nonspecies * 100 / nmags
cat(perct)
```

**MAGs without species-level annotation**
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(MAGs_total = n())
genome_metadata %>%
  filter(species=="") %>%
  group_by(phylum) %>%
  summarize(MAGs_nospecies = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(species_annotated = MAGs_total - MAGs_nospecies) %>%
  mutate(Percentage_nospecies = 100 * MAGs_nospecies / MAGs_total) %>%
  mutate(Percentage_species = 100 - 100 * MAGs_nospecies / MAGs_total) %>%
  tt()
```

### Summary table
```{r single_sp_meta, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

single_sp <- data.frame(Bat = character(), Single_species = numeric())

table_upset_analysis <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample")

unique_all <- table_upset_analysis %>%
  filter(rowSums(across(Eb:Pk)) == 1)

single_sp <- rbind(single_sp, data.frame(Bat = "Total", Single_species = nrow(unique_all)))
  
for (bat in bats) {
  unique <- table_upset_analysis %>%
    filter(rowSums(across(Eb:Pk)) == 1) %>%
    select(all_of(bat)) %>%
    filter(. > 0) %>%
    nrow()
  
  single_sp <- rbind(single_sp, data.frame(Bat = bat, Single_species = unique))
}
```

```{r single_ind_meta, comment="", message=FALSE, warning=FALSE}
single_ind <- data.frame(Bat = character(), Single_individual = numeric())

freq_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("asv")

singleton_filt <- freq_table %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(-asv))) %>% 
  filter(row_sum == 1) %>%
  column_to_rownames(var = "asv")

single_ind <- rbind(single_ind, data.frame(
  Bat = "Total",
  Single_individual = nrow(singleton_filt) 
))

for (bat in bats) {
  single_filt <- singleton_filt %>%
    select(bat) %>%
    filter(. == 1)
  
  single_ind <- rbind(single_ind, data.frame(
    Bat = bat,
    Single_individual = nrow(single_filt)
  ))
}

```


```{r}
summary_table <- total_mags %>%
  left_join(., no_annotation, by = "Bat") %>%
  left_join(., single_ind, by = "Bat") %>%
  left_join(., single_sp, by = "Bat")
summary_table %>% 
  tt()
```

```{r}
summary_table %>%
  select(-Phylum, -Family, -Genus) %>%
  rowwise() %>%
  mutate(Mag_perc=round(MAGs*100/135, 2))%>%
  mutate(No_genus_perc = round(No_genus * 100 / MAGs, 2)) %>%
  mutate(No_species_perc = round(No_species * 100 / MAGs, 2)) %>%
  mutate(Single_individual_perc = round(Single_individual * 100 / MAGs, 2)) %>%
  mutate(Single_species_perc = round(Single_species * 100 / MAGs, 2)) %>%
  mutate(Single_individual_per_Single_species = round(Single_individual * 100 /
           Single_species, 2)) %>%
  select(1,7:12) %>% 
  tt()
```
### Read fractions

```{r microbial_fraction_rel}
microbial_fraction %>%
  select(sample,lowqual_bases,host_bases,unmapped_bases,mapped_bases) %>% 
	pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
	group_by(sample) %>%
  mutate(value = value / sum(value)) %>% 
  ungroup() %>%
	mutate(fraction = factor(fraction, levels = c("lowqual_bases","host_bases","unmapped_bases","mapped_bases"))) %>%
  inner_join(sample_metadata,by="sample") %>% 
	ggplot(., aes(x = sample, y = value, fill=fraction)) +
  geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(name="Sequence type",
                    breaks=c("lowqual_bases","host_bases","unmapped_bases","mapped_bases"),
                    labels=c("Low quality","Mapped to host","Unmapped","Mapped to MAGs"),
                    values=c("#CCCCCC", "#bcdee1", "#d8b8a3","#93655c"))+
  facet_grid(. ~ factor(Species, labels=c("Eb" = "Cnephaeus bottae", "Ha" = "Hypsugo ariel", "Pk" = "Pipistrellus kuhlii")), scales = "free")+
  labs(x = "Samples", y = "Amount of data (GB)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),
        legend.position = "bottom")
```
```{r microbial_fraction_ranges}
microbial_fraction %>%
  mutate(host_fraction=host_bases/(lowqual_bases+host_bases+unmapped_bases+mapped_bases)*100) %>% 
  mutate(mapped_fraction=mapped_bases/(lowqual_bases+host_bases+unmapped_bases+mapped_bases)*100) %>% 
  select(sample,host_fraction,mapped_fraction) %>% 
	pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
  group_by(fraction) %>% 
  summarise(min=min(value),max=max(value)) %>% 
  tt()
```

### Estimated vs recovered proportion

```{r gut_data_estimations_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
microbial_fraction %>% 
	mutate(mapped_microbial_fraction = ifelse(estimated_microbial_fraction == 0, 0, mapped_microbial_fraction)) %>% 
	mutate(estimated_microbial_fraction = ifelse(estimated_microbial_fraction == 0, NA, estimated_microbial_fraction)) %>% 
	mutate(estimated_microbial_fraction = ifelse(estimated_microbial_fraction < mapped_microbial_fraction, NA, estimated_microbial_fraction)) %>% 
	mutate(estimated_microbial_fraction = ifelse(estimated_microbial_fraction > 100, 100, estimated_microbial_fraction)) %>%
  select(sample,estimated_microbial_fraction,mapped_microbial_fraction) %>% 
	pivot_longer(!sample, names_to = "proportion", values_to = "value") %>%
	mutate(proportion = factor(proportion, levels = c("mapped_microbial_fraction","estimated_microbial_fraction"))) %>%
  inner_join(sample_metadata,by="sample") %>%
	ggplot(., aes(x = value, y = sample, color=proportion)) +
  geom_line(aes(group = sample), color = "#f8a538") +
  geom_point() +
  scale_color_manual(name="Proportion",
                    breaks=c("mapped_microbial_fraction","estimated_microbial_fraction"),
                    labels=c("Recovered","Estimated"),
                    values=c("#52e1e8", "#876b53"))+
  facet_grid(rows = vars(Species), space = "free", scales = "free", 
           labeller = labeller(Species = c(Eb = "Cnephaeus bottae", Ha = "Hypsugo ariel", Pk = "Pipistrellus kuhlii")))+  
  theme_classic() +
  labs(y = "Samples", x = "Prokaryotic fraction (%)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "right")

```

### Domain-adjusted mapping rate

```{r gut_data_estimations_plot_do, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
microbial_fraction %>% 
  mutate(damr=ifelse(mapped_microbial_fraction>estimated_microbial_fraction,100,mapped_microbial_fraction/estimated_microbial_fraction*100)) %>% 
  inner_join(sample_metadata,by="sample") %>% 
  summarise(mean=mean(damr),sd=sd(damr)) %>% 
  tt()

microbial_fraction %>%
  mutate(damr=ifelse(mapped_microbial_fraction>estimated_microbial_fraction,100,mapped_microbial_fraction/estimated_microbial_fraction*100)) %>%
  inner_join(sample_metadata,by="sample") %>%
  mutate(Species = factor(Species, levels = c("Eb", "Ha", "Pk"), labels = c("Cnephaeus bottae", "Hypsugo ariel", "Pipistrellus kuhlii"))) %>% 
  group_by(Species) %>% 
  summarise(mean=mean(damr),sd=sd(damr)) %>% 
  tt()

microbial_fraction %>% 
  mutate(damr=ifelse(mapped_microbial_fraction>estimated_microbial_fraction,100,mapped_microbial_fraction/estimated_microbial_fraction*100)) %>% 
  inner_join(sample_metadata,by="sample") %>% 
  kruskal.test(damr ~ Species)
```

## Amplicon
### Taxonomy overview 
```{r div_load_data_ampli, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```

**Number of ASV**
```{r}
genome_metadata %>% 
  nrow() %>% 
  cat()
```

**Number of phyla**
```{r}
genome_metadata %>% 
  distinct(phylum) %>% 
  nrow() %>% 
  cat()
```

**Number of Archaea phyla**
```{r arch_phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

```{r arch_names, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() 
```

**Number of Bacteria phyla**
```{r bact_phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

#### Phylum level
```{r taxonomy_barplot_ampli, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 6),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 2))
```

**Grouping low-abundance bacteria**
```{r group_plot_metab, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
p1 <- genome_counts_filt %>%
  mutate_at(vars(-genome), ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>%
group_by(sample, phylum) %>%
  mutate(total_abundance = sum(count)) %>%
  ungroup() %>%
  mutate(phylum = if_else(total_abundance < 0.01, "Other", phylum)) %>%
  group_by(sample, phylum, Species) %>%
  summarise(count = sum(count), .groups = "drop") %>%
   ggplot(aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = c(phylum_colors, "Other" = "grey50"),
                    drop = FALSE)+
  facet_nested(~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 6),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 2))

#ggsave("community_plot_grouped_standard.pdf", plot = p1, width = 12, height = 6)
p1
```

**Phylum relative abundances**
```{r taxonomy_phylum_summary_ampli1, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, Species) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_phylum_summary_ampli2, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    Total_mean = mean(relabun * 100, na.rm = T),
    Total_sd = sd(relabun * 100, na.rm = T),
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(phylum, Total, Cnephaeus, Hypsugo, Pipistrellus) %>% 
  tt()
```

**Number of phyla in each bat species**
```{r}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(Species = factor(Species, levels = c("Eb", "Ha", "Pk"), labels = c("Cnephaeus bottae", "Hypsugo ariel", "Pipistrellus kuhlii"))) %>% 
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,domain,phylum,Species) %>%
  summarise(relabun=sum(count))

phylum_summary %>% 
  filter(relabun > 0) %>% 
  group_by(Species,domain) %>% 
  summarise(n_phyla = n_distinct(phylum)) %>% 
  tt()
```

#### Family level

**Percentage of families in each group**
```{r taxonomy_family_summary_ampli, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  group_by(sample, phylum, family, Species) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_family_summary_ampli1, warning=FALSE, comments="", message=FALSE}
family_summary %>%
  group_by(phylum,family) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Pk_mean) %>%
  dplyr::select(family, Cnephaeus, Hypsugo, Pipistrellus) %>%
  tt()
```

#### Genus level

**Percetage of genera in each species**
```{r taxonomy_genus_summary_ampli, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, genus, Species) %>%
  summarise(relabun = sum(count))

genus_summary %>%
  group_by(genus) %>%
  summarise(
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Pk_mean) %>%
  dplyr::select(genus, Cnephaeus, Hypsugo, Pipistrellus) %>%
  left_join(., genome_metadata[c(3, 7)] %>% unique(), by = join_by(genus == genus)) %>% 
  tt()
```

### ASVs

**Number of ASVs and distinct taxonomy**
```{r total_ASv_ampli, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

total_asvs <- data.frame(
  Bat = character(),
  MAGs = numeric(),
  Phylum = numeric(),
  Family = numeric(),
  Genus = numeric()
)

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome")) %>% 
  filter(domain=="Bacteria")

phylum <- preabs_table %>%
  distinct(phylum)

family <- preabs_table %>%
  distinct(phylum, class, order, family)

genus <- preabs_table %>%
  distinct(phylum, class, order, family, genus)

total_asvs <- rbind(
  total_asvs,
  data.frame(
    Bat = "Total",
    ASVs = nrow(preabs_table),
    Phylum = nrow(phylum),
    Family = nrow(family),
    Genus = nrow(genus)
  )
)

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  phylum <- preabs_table %>%
    select({{bat}}, phylum) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum)
  
  family <- preabs_table %>%
    select({{bat}}, phylum, class, order, family) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family, genus)
  
  total_asvs <- rbind(
    total_asvs,
    data.frame(
      Bat = bat,
      ASVs = nrow(number),
      Phylum = nrow(phylum),
      Family = nrow(family),
      Genus = nrow(genus)
    )
  )
}
```

```{r total_annot_ampli, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

no_annotation <- data.frame(Bat = character(),
                            No_genus = numeric(),
                            No_species = numeric())

preabs_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata, by = join_by("genome" == "genome"))%>% 
  filter(domain=="Bacteria")

genus <- preabs_table  %>%
  filter(is.na(genus))

species <- preabs_table  %>%
  filter(is.na(species))

no_annotation <- rbind(no_annotation,
                       data.frame(
                         Bat = "Total",
                         No_genus = nrow(genus),
                         No_species = nrow(species)
                       ))

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(genus))
  
  species <- preabs_table %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(species))
  
  no_annotation <- rbind(no_annotation,
                         data.frame(
                           Bat = bat,
                           No_genus = nrow(genus),
                           No_species = nrow(species)
                         ))
}
```

**Total percentage of ASVs without genus-level annotation**
```{r gen_percet_ampli, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>% 
  filter(domain=="Bacteria") %>%
  filter(is.na(genus)) %>% 
  summarize(ASV_nogenera = n()) %>%
  pull()
nasvs <- total_asvs %>%
  filter(Bat == "Total") %>%
  select(ASVs) %>%
  pull()
perct <- nongenera * 100 / nasvs
cat(perct)
```

**Percentage of ASVs without genus-level annotation by phylum**
```{r nongene_phylum_gut_ampli, comment="", message=FALSE, warning=FALSE}
total_asv_phylum <- genome_metadata %>% 
  filter(domain=="Bacteria") %>%
  group_by(phylum) %>%
  summarize(Total_ASVs = n())
genome_metadata %>% 
  filter(domain=="Bacteria") %>%
  filter(is.na(genus)) %>% 
  group_by(phylum) %>%
  summarize(ASVs_nogenus = n()) %>%
  left_join(total_asv_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Percentage_nogenus = 100 * ASVs_nogenus / Total_ASVs) %>%
  tt()
```

**Number of bacterial species**
```{r phyla_ampli, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(species) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Total percentage of MAGs without species-level annotation**
```{r sp_percet_ampli, comment="", message=FALSE, warning=FALSE}
nonspecies <- genome_metadata %>% 
  filter(domain=="Bacteria")%>%
  filter(is.na(species)) %>%
  summarize(ASV_nospecies = n()) %>%
  pull()
perct <- nonspecies * 100 / nasvs
cat(perct)
```

**ASVs without species-level annotation**
```{r nonspe_phylum_ampli, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>% 
  filter(domain=="Bacteria") %>%
  group_by(phylum) %>%
  summarize(ASVs_total = n())

genome_metadata %>% 
  filter(domain=="Bacteria")%>%
  filter(is.na(species)) %>%
  group_by(phylum) %>%
  summarize(ASVs_nospecies = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Species_annotated = ASVs_total - ASVs_nospecies) %>%
  mutate(Percentage_nospecies = 100 * ASVs_nospecies / ASVs_total) %>%
  mutate(Percentage_species = 100 - 100 * ASVs_nospecies / ASVs_total) %>%
  tt()
```

### Summary table

```{r single_sp_ampli, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

single_sp <- data.frame(Bat = character(), Single_species = numeric())

bacteria <- genome_metadata %>%
 filter(domain=="Bacteria")

table_upset_analysis <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  filter(genome %in% bacteria$genome) %>% 
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample")

unique_all <- table_upset_analysis %>%
  filter(rowSums(across(Eb:Pk)) == 1)

single_sp <- rbind(single_sp, data.frame(Bat = "Total",
                                         Single_species = nrow(unique_all)))
  
for (bat in bats) {
  unique <- table_upset_analysis %>%
    filter(rowSums(across(Eb:Pk)) == 1) %>%
    column_to_rownames(., "sample") %>% 
    select(bat) %>%
    filter(. > 0) %>%
    nrow()
  single_sp <- rbind(single_sp, data.frame(Bat = bat, Single_species = unique))
}
```

```{r single_ind_ampli, comment="", message=FALSE, warning=FALSE}
single_ind <- data.frame(Bat = character(), Single_individual = numeric())

freq_table <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  filter(genome %in% bacteria$genome) %>% 
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("asv")

singleton_filt <- freq_table %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(-asv))) %>%
  filter(row_sum == 1) %>%
  column_to_rownames(var = "asv")  %>%
  filter(row_sum == 1)

single_ind <- rbind(single_ind, data.frame(
  Bat = "Total",
  Single_individual = nrow(singleton_filt)
))
     
for (bat in bats) {
  singleton_filt <- freq_table %>%
    rowwise() %>%
    mutate(row_sum = sum(c_across(-asv))) %>%
    filter(row_sum == 1) %>%
    column_to_rownames(var = "asv")  %>%
    select(bat) %>%
    filter(. == 1)
  
  single_ind <- rbind(single_ind, data.frame(
    Bat = bat,
    Single_individual = nrow(singleton_filt) 
  ))
}
```

```{r merge_ampli, comment="", message=FALSE, warning=FALSE}
summary_table_ampli <- total_asvs %>%
  left_join(., no_annotation, by = "Bat") %>%
  left_join(., single_ind, by = "Bat") %>%
  left_join(., single_sp, by = "Bat") %>%
  mutate(Bat = recode(Bat, 
                      Eb = "Cnephaeus bottae",
                      Ha = "Hypsugo ariel",
                      Pk = "Pipistrellus kuhlii"))
summary_table_ampli %>% 
  tt()
```

```{r summary_table_ampli, comment="", message=FALSE, warning=FALSE}
total_asv <- genome_metadata %>% 
  filter(domain=="Bacteria") %>% 
  nrow()
summary_table_ampli %>% 
  select(-Phylum,-Family, -Genus) %>% 
  rowwise() %>% 
  mutate(ASV_perc=round(ASVs*100/total_asv, 2))%>% 
  mutate(No_genus_perc=round(No_genus*100/ASVs, 2))%>% 
  mutate(No_species_perc=round(No_species*100/ASVs, 2)) %>% 
  mutate(Single_individual_perc=round(Single_individual*100/ASVs, 2))%>% 
  mutate(Single_species_perc=round(Single_species*100/ASVs, 2)) %>% 
  mutate(Single_individual_per_unique=round(Single_individual*100/Single_species, 2)) %>% 
  select(1,7:12) %>% 
  tt()
```

### Archaea

**Number of ASV**
```{r}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  nrow() %>% 
  cat()
```

**Archaeal ASVs present in more than one individual**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>%
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0)))%>%
  mutate(count = rowSums(across(-genome))) %>% 
  select(genome, count) %>% 
  filter(count>1)
```

**Presence of archaeal ASVs**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  select(
    genome,
    where(~ sum(. > 0) > 0)
  ) %>%
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>% 
  select(-genome) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(archaea=1) %>% 
  rownames_to_column("sample") %>% 
  left_join(., sample_metadata[c(1,6)], by="sample") %>% 
  tt()
```

**Number of archaeal ASVs in each species**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  select(
    genome,
    where(~ sum(. > 0) > 0)
  ) %>%
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>% 
  select(-genome) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(archaea=1) %>% 
  rownames_to_column("sample") %>% 
  left_join(., sample_metadata[c(1,6)], by="sample") %>% 
  group_by(species_name) %>% 
  summarise(asv=sum(archaea)) %>% 
  tt()
```

**Number of individuals with archaeal ASVs**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  select(
    genome,
    where(~ sum(. > 0) > 0)
  ) %>%
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>% 
  select(-genome) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(archaea=1) %>% 
  rownames_to_column("sample") %>% 
  left_join(., sample_metadata[c(1,6)], by="sample") %>% 
  group_by(species_name) %>% 
  summarise(indiv=n()) %>% 
  tt()
```

**Mean presence of archaeal ASVs**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>% 
  select(-genome) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(archaea=1) %>% 
  rownames_to_column("sample") %>% 
  left_join(., sample_metadata[c(1,6)], by="sample") %>% 
  summarise(mean=mean(archaea), sd=sd(archaea)) %>% 
  tt()
```

**Mean presence of archaeal ASVs per species**
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>% 
  select(-genome) %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(archaea=1) %>% 
  rownames_to_column("sample") %>% 
  left_join(., sample_metadata[c(1,6)], by="sample") %>% 
  group_by(species_name) %>% 
  summarise(mean=mean(archaea), sd=sd(archaea)) %>% 
  tt()
```

```{r}
genome_archaea %>% 
  distinct(genus) %>% 
  tt()
```
```{r}
genome_counts_filt %>% 
  filter(genome %in% genome_archaea$genome) %>% 
  select(
    genome,
    where(~ sum(. > 0) > 0)
  ) %>%
  mutate(across(-genome, ~ ifelse(. > 0, 1, 0))) %>%
  mutate(count = rowSums(across(-genome))) %>% 
  select(genome, count) %>% 
  arrange(-count)  %>% 
  tt()
```

