# Alpha diversity
```{r knitr_options_wild, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(1703)
```

## Metagenomics
```{r load_data_alpha_metagenomics, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
species_color <- c("#e5bd5b", "#6b7398", "#76b183")

sample_metadata <- sample_metadata %>%
  select(1,2,7,8) %>%
  mutate(method="Metagenomics")
```

### MAG level
```{r alpha_div, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

# Merge all metrics
alpha_div_meta <- richness %>%
  full_join(neutral, by = join_by(sample == sample))
```

```{r}
microbial_fraction_mapped <- microbial_fraction %>% 
  select(sample, mapped_microbial_fraction) %>% 
  right_join(alpha_div_meta, by="sample") %>% 
  left_join(sample_metadata, by="sample")

microbial_fraction_mapped %>% 
  summarise(
    Total_mean = mean(mapped_microbial_fraction, na.rm = T),
    Total_sd = sd(mapped_microbial_fraction, na.rm = T),
    Eb_mean = mean(mapped_microbial_fraction[Species == "Eb"], na.rm = T),
    Eb_sd = sd(mapped_microbial_fraction[Species == "Eb"], na.rm = T),
    Ha_mean = mean(mapped_microbial_fraction[Species == "Ha"], na.rm = T),
    Ha_sd = sd(mapped_microbial_fraction[Species == "Ha"], na.rm = T),
    Pk_mean = mean(mapped_microbial_fraction[Species == "Pk"], na.rm = T),
    Pk_sd = sd(mapped_microbial_fraction[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(
                Total,
                Cnephaeus,
                Hypsugo,
                Pipistrellus) %>%
  tt()
```
```{r}
  kruskal.test(mapped_microbial_fraction ~ Species, data = microbial_fraction_mapped)
```

Correlation between diversity and estimate microbial fraction
```{r}
cor.test(microbial_fraction_mapped$mapped_microbial_fraction, microbial_fraction_mapped$richness, method = "spearman")
cor.test(microbial_fraction_mapped$mapped_microbial_fraction, microbial_fraction_mapped$neutral, method = "spearman")

# Histogram
hist(microbial_fraction_mapped$mapped_microbial_fraction, main = "Read Fraction", xlab = "read_fraction")
hist(microbial_fraction_mapped$richness, main = "Richness", xlab = "richness")

# QQ plots
qqnorm(microbial_fraction_mapped$mapped_microbial_fraction); qqline(microbial_fraction_mapped$mapped_microbial_fraction)
qqnorm(microbial_fraction_mapped$richness); qqline(microbial_fraction_mapped$richness)

shapiro.test(microbial_fraction_mapped$mapped_microbial_fraction)
shapiro.test(microbial_fraction_mapped$richness)
```

```{r fig_fraction_mergedmeta, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
p1 <- ggplot(microbial_fraction_mapped, aes(x = mapped_microbial_fraction, y = richness)) +
  geom_point(size = 3, color = "#2E86AB") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Richness")

p2 <- ggplot(microbial_fraction_mapped, aes(x = mapped_microbial_fraction, y = neutral)) +
  geom_point(size = 3, color = "#F39C12") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Neutral Alpha Diversity")

p1 + p2
```
```{r alpha_meta_table, comment="", message=FALSE, warning=FALSE}
metagenome_alpha_div <- alpha_div_meta %>%
  select(1:3) %>% 
  left_join(sample_metadata, by = join_by(sample == sample))
```

```{r save_alpha_meta, comment="", message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(metagenome_alpha_div, "resources/metagenomics/metagenome_alpha_div.rds")
```

**Diversity values at MAG level**
```{r alpha_div_diets_summary_all_metagenomics, comment="", message=FALSE, warning=FALSE}
alpha_div_meta %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_general, comment="", message=FALSE, warning=FALSE}
Model_richness <- MASS::glm.nb(richness ~ Species, data = metagenome_alpha_div,trace=TRUE)
summary(Model_richness)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

**Neutral**
```{r neutral_MM_general, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = metagenome_alpha_div)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

### At different taxonomic levels
```{r taxo, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")

genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

#### Family level
```{r alpha_family_metag_indiv, comment="", message=FALSE, warning=FALSE}
richness_f_metag <- family_level_agg %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_f_metag <- family_level_agg %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_f_metag <- richness_f_metag %>%
  full_join(neutral_f_metag, by = join_by(sample == sample))

alpha_div_f_metag_data <- alpha_div_f_metag %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_fam_summary_all_metagenomics, comment="", message=FALSE, warning=FALSE}
alpha_div_f_metag %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_general_metag_indiv, comment="", message=FALSE, warning=FALSE}
Model_richness <- MASS::glm.nb(richness ~ Species, data = alpha_div_f_metag_data,trace=TRUE)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

**Neutral**
```{r neutral_MM_general_metag_indiv, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_f_metag_data)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

#### Phylum level
```{r alpha_phylum_metag_indiv, comment="", message=FALSE, warning=FALSE}
richness_p_metag <- phylum_level_agg %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_p_metag <- phylum_level_agg %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_phylum_metag <- richness_p_metag %>%
  full_join(neutral_p_metag, by = join_by(sample == sample))

alpha_div_phylum_metag_data <- alpha_div_phylum_metag %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_phy_summary_all_metagenomics, comment="", message=FALSE, warning=FALSE}
alpha_div_phylum_metag %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_phylum_metag_indiv, comment="", message=FALSE, warning=FALSE}
Model_richness_p <- MASS::glm.nb(richness ~ Species, data = alpha_div_phylum_metag_data,trace=TRUE)
anova(Model_richness_p)
emmeans(Model_richness_p, pairwise ~ Species)
```

**Neutral**
```{r neutral_MM_phylum_metag_indiv, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_phylum_metag_data)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

## Amplicon
```{r div_load_data, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
sample_metadata <- sample_metadata %>%
  select(1,9,10,13) %>%
  mutate(method="Metabarcoding_standard")

no_archaea <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,2)], by="genome") %>% 
  filter(domain!="Archaea") %>% 
  select(-domain)
```

### ASV level
```{r filter_copy2, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

amplicon_alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample))
```

**Correlation between diversity and estimate microbial fraction**
```{r corr,comment="", message=FALSE, warning=FALSE}
fraction_ampli <- read_csv("resources/metagenomics/microbial_fraction_estimation.csv") %>% 
  mutate(sample = str_remove(sample, "_M_1")) %>% 
  select(sample, read_fraction) %>% 
  right_join(amplicon_alpha_div, by="sample") %>% 
  left_join(sample_metadata, by="sample")

cor.test(fraction_ampli$read_fraction, fraction_ampli$richness, method = "spearman")
cor.test(fraction_ampli$read_fraction, fraction_ampli$neutral, method = "spearman")
```

```{r fraction_ampli_fig, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
p1_ampli <- ggplot(fraction_ampli, aes(x = read_fraction, y = richness)) +
  geom_point(size = 3, color = "#2E86AB") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Richness", title= "Correlation between estimated microbial fraction and diversity")

p2_ampli <- ggplot(fraction_ampli, aes(x = read_fraction, y = neutral)) +
  geom_point(size = 3, color = "#F39C12") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Neutral Alpha Diversity")

p1_ampli + p2_ampli 
```

**Diversity values**
```{r alpha_div_diets_summary_all_amplicon, comment="",echo=FALSE, message=FALSE, warning=FALSE}
amplicon_alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

```{r alpha_amp_table, comment="", message=FALSE, warning=FALSE}
amplicon_alpha_div_data <- amplicon_alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```
```{r save_alpha_amp, comment="", message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(amplicon_alpha_div_data, "resources/amplicon/amplicon_alpha_div.rds")
```

**Richness**
```{r rich_MM_ampli_species_sp, comment="", message=FALSE, warning=FALSE}
Model_richness_ampli <- MASS::glm.nb(richness ~ Species, data = amplicon_alpha_div_data,trace=TRUE)
anova(Model_richness_ampli)
```

**Neutral**
```{r neutral_MM_ampli_species_sp, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = amplicon_alpha_div_data)
anova(Model_neutral)
```

### At different taxonomic level
```{r}
load("resources/amplicon/data_standard_filt.Rdata")

sample_metadata_amplicon <- sample_metadata[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding_standard")

genome_metadata_ampli <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

#### Family level
```{r alpha_family_indiv_ampli, comment="", message=FALSE, warning=FALSE}
richness_f_ampli <- family_level_agg_ampli %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_f_ampli <- family_level_agg_ampli %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_f_ampli <- richness_f_ampli %>%
  full_join(neutral_f_ampli, by = join_by(sample == sample))

alpha_div_f_ampli_data <- alpha_div_f_ampli %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_fam_summary_all_ampli, comment="", message=FALSE, warning=FALSE}
alpha_div_f_ampli %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_family_indiv_ampli, comment="", message=FALSE, warning=FALSE}
Model_richness_ampli_f <- MASS::glm.nb(richness ~ Species, data = alpha_div_f_ampli_data,trace=TRUE)
anova(Model_richness_ampli_f)
```

**Neutral**
```{r neutral_MM_family_indiv_ampli, comment="", message=FALSE, warning=FALSE}
Model_neutral_ampli_f <- lm(formula = neutral ~ Species, data = alpha_div_f_ampli_data)
anova(Model_neutral_ampli_f)
```

#### Phylum level
```{r alpha_phylum_indiv_ampli, comment="", message=FALSE, warning=FALSE}
richness_p_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_p_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_phylum_ampli <- richness_p_ampli %>%
  full_join(neutral_p_ampli, by = join_by(sample == sample))

alpha_div_phylum_ampli_data <- alpha_div_phylum_ampli %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_phy_summary_all_ampli, comment="", message=FALSE, warning=FALSE}
alpha_div_phylum_ampli %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_phylum_indiv_ampli, comment="", message=FALSE, warning=FALSE}
Model_richness_ampli_p <- MASS::glm.nb(richness ~ Species, data = alpha_div_phylum_ampli_data,trace=TRUE)
anova(Model_richness_ampli_p)
```

**Neutral**
```{r neutral_MM_phylum_indiv_ampli, comment="", message=FALSE, warning=FALSE}
Model_neutral_ampli_p <- lm(formula = neutral ~ Species, data = alpha_div_phylum_ampli_data)
anova(Model_neutral_ampli_p)
```

## Comparison
### At family and phylum level

```{r}
load("resources/amplicon/data_standard_filt.Rdata")

sample_metadata_amplicon <- sample_metadata[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding_standard")

genome_metadata_ampli <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

```{r merging, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")

genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

```{r}
merged_table <- full_join(family_level_agg_ampli,family_level_agg, 
          by = c("family"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

merged_table_phylum <- full_join(phylum_level_agg_ampli,phylum_level_agg, 
          by = c("phylum"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

sample_shot <- sample_metadata %>% 
  select(-c(Habitat,Longitude, Latitude, Region))%>%
  mutate(method="Metagenomics")

all_samples <- rbind(sample_shot, sample_metadata_amplicon)
```

```{r save_alpha, comment="", message=FALSE,echo=FALSE,warning=FALSE, eval=FALSE}
save(merged_table, 
     merged_table_phylum,
     all_samples,
     file = "resources/alpha_together.Rdata")
```

```{r alpha_family_comp, comment="", message=FALSE, warning=FALSE}
richness <- merged_table %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- merged_table %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_family <- richness %>%
  full_join(neutral, by = join_by(sample == sample))

alpha_div_family_data <- alpha_div_family %>%
  left_join(all_samples, by = join_by(sample == sample)) %>% 
  mutate(sampleid = str_remove(sample, "_ampli"))
```

#### Family level
```{r fig_family_comp, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=16, fig.fullwidth=TRUE}
alpha_div_family_data %>%
  select(-sample,-sampleid, -Life_stage, -Sex) %>%
  pivot_longer(
    cols = -c(method, Species),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("richness", "neutral"),
      labels = c("Richness", "Neutral")
    ),
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )
  ) %>%  
  ggplot(aes(y = value, x = method, color = Species, fill = Species)) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
    size = 1,
    alpha = 0.8,
  ) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  scale_fill_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = 0.1, color = "grey"),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
    strip.text = element_text(size = 14, lineheight = 0.6)
  ) +
  labs(
    x = "Methodological approach",
    y = "Alpha diversity at family level",
    color = "Species"
  )
```

**Richness**
```{r rich_MM_ampli_species_family, comment="", message=FALSE, warning=FALSE}
Model_richness_stand <- glmer.nb(richness ~ Species*method+(1|sampleid), data = alpha_div_family_data)
summary(Model_richness_stand)
anova(Model_richness_stand)
emmeans(Model_richness_stand, pairwise ~ Species)
emmeans(Model_richness_stand, pairwise ~ method | Species)
```

**Neutral**
```{r neutral_MM_ampli_species_family, comment="", message=FALSE, warning=FALSE}
Modelq1 <- lme(fixed = neutral ~ Species*method, data = alpha_div_family_data,
               random = ~ 1 | sampleid)
summary(Modelq1)
anova(Modelq1)
emmeans(Modelq1, pairwise ~ Species)
emmeans(Modelq1, pairwise ~ method | Species)
```

#### Phylum level
```{r alpha_phylum_ampli, comment="", message=FALSE, warning=FALSE}
richness_phylum <- merged_table_phylum %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_phylum <- merged_table_phylum %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_phylum <- richness_phylum %>%
  full_join(neutral_phylum, by = join_by(sample == sample))
alpha_div_phylum_data <- alpha_div_phylum %>%
  left_join(all_samples, by = join_by(sample == sample)) %>% 
  mutate(sampleid = str_remove(sample, "_ampli"))
```

```{r gif_phylum_amp, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=16, fig.fullwidth=TRUE}
alpha_div_phylum_data %>%
  select(-sample, -sampleid, -Life_stage, -Sex) %>%
  pivot_longer(
    cols = -c(method, Species),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("richness", "neutral"),
      labels = c("Richness", "Neutral")
    ),
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )
  ) %>%   
  ggplot(aes(y = value, x = method, color = Species, fill = Species)) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
    size = 1,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  scale_fill_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = 0.1, color = "grey"),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
    strip.text = element_text(size = 14, lineheight = 0.6)
  ) +
  labs(
    x = "Methodological approach",
    y = "Alpha diversity at phylum level",
    color = "Species"
  )
```

**Richness**
```{r rich_alpha_phylum_comp, comment="", message=FALSE, warning=FALSE}
Model_richness_random <- glmer.nb(richness ~ Species*method+(1|sampleid), data = alpha_div_phylum_data)
summary(Model_richness_random)
anova(Model_richness_random)
emmeans(Model_richness_random, pairwise ~ Species)
emmeans(Model_richness_random, pairwise ~ method | Species)
```

**Neutral**
```{r neutral_alpha_phylum_comp, comment="", message=FALSE, warning=FALSE}
Modelq1_random <- lme(fixed = neutral ~ Species*method, data = alpha_div_phylum_data,
               random = ~ 1 | sampleid)
summary(Modelq1_random)
anova(Modelq1_random)
emmeans(Modelq1_random, pairwise ~ Species)
emmeans(Modelq1_random, pairwise ~ method | Species)
```