# Beta diversity

```{r knitr_options_beta, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(24022025)
```

## Metagenomics
```{r load_data_beta}
load("resources/metagenomics/data_filtered.Rdata")
species_color <- c("#e5bd5b", "#6b7398", "#76b183")
```

***Identifying outliers***
```{r beta_div_nmds_neutral_plot_out, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(
    x = NMDS1,
    y = NMDS2,
    color = Species
  )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  theme(axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical") +
  labs(color = "Species")+geom_text_repel(aes(label = sample), size=3)
```

```{r}
load("resources/metagenomics/beta_filtered.Rdata")
sample_metadata <- sample_metadata %>%
  filter(!sample %in% c("H06", "H20"))
genome_counts_filt <- genome_counts_filt %>%
  select(-H06, -H20) %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0))
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

```{r save_beta_meta, comment="", message=FALSE,echo=FALSE,warning=FALSE, eval=FALSE}
save(beta_q0n, 
     beta_q1n,
     file = "resources/metagenomics/beta_filtered.Rdata")
```

### MAG level

**Richness**
```{r richness_metag, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q0n$S, sample_metadata$Species) %>% permutest(.)
adonis2(
  beta_q0n$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_richness_metagenomics, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q0n$S, sample_metadata$Species, perm = 999)
```

**Neutral**
```{r permanova_neutral_metag, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q1n$S, sample_metadata$Species) %>% permutest(.)

adonis2(
  beta_q1n$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_neutral_metagenomics, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q1n$S, sample_metadata$Species, perm = 999)
```

#### NMDS

**Richness**
```{r beta_div_nmds_richness_plot, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

**Neutral**
```{r beta_div_nmds_neutral_plot, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(
    x = NMDS1,
    y = NMDS2,
    color = Species
  )) +
  geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

### At different taxonomic level
```{r }
genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

#### Family
```{r pairwise_richness_family_metag, comment="", message=FALSE, warning=FALSE}
beta_q0n_shot <- family_level_agg %>%
  select(-H09) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 0)

adonis2(
  beta_q0n_shot$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n_shot$S)) %>%
    arrange(match(sample, labels(beta_q0n_shot$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n_shot$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n_shot$S)) %>%
    arrange(match(sample, labels(beta_q0n_shot$S))) %>%
    pull(Species),
  perm = 999
)
```
```{r pairwise_neutral_family_metag, comment="", message=FALSE, warning=FALSE}
beta_q1n_shot <- family_level_agg %>%
  select(-H19) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 1)

adonis2(
  beta_q1n_shot$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q1n_shot$S)) %>%
    arrange(match(sample, labels(beta_q1n_shot$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q1n_shot$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q1n_shot$S)) %>%
    arrange(match(sample, labels(beta_q1n_shot$S))) %>%
    pull(Species),
  perm = 999
)
```

```{r beta_div1_both_fig7, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_shot$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )) %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +  
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

#### Phylum
```{r pairwise_richness_phylum_metag, comment="", message=FALSE, warning=FALSE}
beta_q0n_P_shot <- phylum_level_agg %>%
  select(-H09) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 0)

adonis2(
  beta_q0n_P_shot$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n_P_shot$S)) %>%
    arrange(match(sample, labels(beta_q0n_P_shot$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n_P_shot$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n_P_shot$S)) %>%
    arrange(match(sample, labels(beta_q0n_P_shot$S))) %>%
    pull(Species),
  perm = 999
)
```
```{r pairwise_neutral_phylum_metag, comment="", message=FALSE, warning=FALSE}
beta_q1n_P_shot <- phylum_level_agg %>%
  select(-H19) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 1)

adonis2(
  beta_q1n_P_shot$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q1n_P_shot$S)) %>%
    arrange(match(sample, labels(beta_q1n_P_shot$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q1n_P_shot$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q1n_P_shot$S)) %>%
    arrange(match(sample, labels(beta_q1n_P_shot$S))) %>%
    pull(Species),
  perm = 999
)
```

## Amplicon
```{r div_load_data_ampli2, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```

```{r beta_div_ampli, comment="", message=FALSE, warning=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

### ASV level

**Richness**
```{r permanova_ampli, comment="", message=FALSE, warning=FALSE}
betadisper(
  beta_q0n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species)
) %>%
  permutest(.)

adonis2(
  beta_q0n$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_rich, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(
  beta_q0n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species),
  perm = 999
)
```

**Neutral**
```{r permanova_neutral_amplicon1, comment="", message=FALSE, warning=FALSE}
betadisper(
  beta_q1n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species)
) %>%
  permutest(.)

adonis2(
  beta_q1n$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```

```{r pairwise_neutral_ampli, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q1n$S, sample_metadata %>% filter(sample %in% labels(beta_q0n$S)) %>% arrange(match(sample, labels(beta_q0n$S))) %>% pull(Species), perm = 999) %>% 
  tt()
```

#### NMDS

**Richness**
```{r beta_div_nmds_richness_plot_ampli, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
 geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

**Neutral**
```{r beta_div_nmds_neutral_plot_ampli, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
   geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")

#ggsave("Figures_ms/beta_neutral_metabarcoding.pdf", plot = p, width = 8, height = 4)
```

### At different taxonomic level
```{r}
sample_metadata_amplicon <- sample_metadata[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding_standard")

genome_metadata_ampli <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

#### Family level
```{r}
beta_q0n_ampli <- family_level_agg_ampli %>%
  column_to_rownames(., "family") %>% 
  hillpair(., q = 0)

adonis2(
  beta_q0n_ampli$S ~ Species,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n_ampli$S)) %>%
    arrange(match(sample, labels(beta_q0n_ampli$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n_ampli$S,
  all_samples %>%
    filter(sample %in% labels(beta_q0n_ampli$S)) %>%
    arrange(match(sample, labels(beta_q0n_ampli$S))) %>%
    pull(Species),
  perm = 999) %>% 
  tt()
```
```{r}
beta_q1n_ampli <- family_level_agg_ampli %>%
  column_to_rownames(., "family") %>% 
  hillpair(., q = 1)

adonis2(
  beta_q1n_ampli$S ~ Species,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_ampli$S)) %>%
    arrange(match(sample, labels(beta_q1n_ampli$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q1n_ampli$S,
  all_samples %>%
    filter(sample %in% labels(beta_q1n_ampli$S)) %>%
    arrange(match(sample, labels(beta_q1n_ampli$S))) %>%
    pull(Species),
  perm = 999) %>% 
  tt()
```
```{r beta_div1_both_fig5, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )) %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 3) +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +  
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

#### Phylum level
```{r}
beta_q0n_P_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 0)

adonis2(
  beta_q0n_P_ampli$S ~ Species,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n_P_ampli$S)) %>%
    arrange(match(sample, labels(beta_q0n_P_ampli$S))),
  permutations = 999) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n_P_ampli$S,
  all_samples %>%
    filter(sample %in% labels(beta_q0n_P_ampli$S)) %>%
    arrange(match(sample, labels(beta_q0n_P_ampli$S))) %>%
    pull(Species),
  perm = 999) %>%
  tt()
```
```{r}
beta_q1n_P_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 1)

adonis2(
  beta_q1n_P_ampli$S ~ Species,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_P_ampli$S)) %>%
    arrange(match(sample, labels(beta_q1n_P_ampli$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q1n_ampli$S,
  all_samples %>%
    filter(sample %in% labels(beta_q1n_P_ampli$S)) %>%
    arrange(match(sample, labels(beta_q1n_P_ampli$S))) %>%
    pull(Species),
  perm = 999)  %>%
  tt()
```

## Comparison 

*Aggregate counts at family or phylum level in both datasets*
```{r}
load("resources/amplicon/data_standard_filt.Rdata")

sample_metadata_amplicon <- sample_metadata[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding_standard")

genome_metadata_ampli <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

```{r }
load("resources/metagenomics/data_filtered.Rdata")

genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

```{r}
merged_table <- full_join(family_level_agg_ampli,family_level_agg, 
          by = c("family"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

merged_table_phylum <- full_join(phylum_level_agg_ampli,phylum_level_agg, 
          by = c("phylum"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


sample_shot <- sample_metadata %>% 
  select(-c(Habitat,Longitude, Latitude, Region))%>%
  mutate(method="Metagenomics")

all_samples <- rbind(sample_shot, sample_metadata_amplicon)
```

```{r beta_div_together_family, comment="", message=FALSE, warning=FALSE}
beta_q0n <- merged_table %>%
  select(-H19, -H20) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 0)

beta_q1n <- merged_table %>%
  select(-H19) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 1)

beta_q0n_phylum <- merged_table_phylum %>%
  select(-H19, -H20) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 0)

beta_q1n_phylum <- merged_table_phylum %>%
  select(-H19) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 1)
```

### Family
**Richness**
```{r beta_div1_both_fig1, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(
    x = "Methodological approach",
    y = "Alpha diversity at phylum level",
    shape = "Methodological approach"
  )#+labs(color = "method") +geom_text_repel(aes(label = sample), size=3)
  
```

```{r permanova_rich_both_fig_family, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q0n$S, all_samples %>% filter(sample %in% labels(beta_q0n$S)) %>% arrange(match(sample, labels(beta_q0n$S))) %>% pull(Species)) %>% permutest(.)

all_samples <- all_samples %>% 
  mutate(id=sample) %>% 
  mutate(id = str_remove(id, "_ampli$"))

adonis2(
  beta_q0n$S ~ Species*method,
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999, 
  strata = all_samples %>% filter(sample %in% labels(beta_q0n$S)) %>% arrange(match(sample,labels(beta_q0n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q0n$S ~ Species*method,
  by = "term",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999, 
  strata = all_samples %>% filter(sample %in% labels(beta_q0n$S)) %>% arrange(match(sample,labels(beta_q0n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q0n$S ~ Species+method,
  by = "margin",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999, 
  strata = all_samples %>% filter(sample %in% labels(beta_q0n$S)) %>% arrange(match(sample,labels(beta_q0n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()
```

**Neutral**
```{r beta_div1_both_fig3_Family, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 3) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(shape = "Methodological approach")
```

```{r permanova_neutral_tog_Family, comment="", message=FALSE, warning=FALSE}
betadisper(
  beta_q1n$S,
  all_samples %>%
    filter(sample %in% labels(beta_q1n$S)) %>%
    arrange(match(sample, labels(beta_q1n$S))) %>%
    pull(method)
) %>%
  permutest(.)

adonis2(
  beta_q1n$S ~ Species*method,
  data = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q1n$S ~ Species*method,
  by = "terms",
  data = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q1n$S ~ Species+method,
  by = "margin",
  data = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n$S)) %>% arrange(match(sample,labels(beta_q1n$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()
```

### Phylum
**Richness**
```{r beta_div1_both_fig_phylum, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q0n_phylum$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +labs(color = "method") + geom_text_repel(aes(label = sample), size=3)
  
```

```{r permanova_rich_both_phylum_fig2, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q0n_phylum$S, all_samples %>% filter(sample %in% labels(beta_q0n_phylum$S)) %>% arrange(match(sample, labels(beta_q0n_phylum$S))) %>% pull(Species)) %>% permutest(.)

adonis2(beta_q0n_phylum$S ~ Species*method, 
        data = all_samples %>% filter(sample %in% labels(beta_q0n_phylum$S)) %>% arrange(match(sample, labels(beta_q0n_phylum$S))), permutations = 999) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q0n_phylum$S ~ Species*method,
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q0n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q0n_phylum$S)) %>% arrange(match(sample,labels(beta_q0n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q0n_phylum$S ~ Species*method,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q0n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q0n_phylum$S)) %>% arrange(match(sample,labels(beta_q0n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q0n_phylum$S ~ Species+method,
  by = "margin",
  data = all_samples %>%
    filter(sample %in% labels(beta_q0n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q0n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q0n_phylum$S)) %>% arrange(match(sample,labels(beta_q0n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()
```

**Neutral**
```{r beta_div1_both_fig3_phylum, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q1n_phylum$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Pk", "Ha"),
      labels = c("C. bottae", "P. kuhlii", "H. ariel")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 3) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(shape = "Methodological approach")
```

```{r permanova_neutral_tog_phylum, comment="", message=FALSE, warning=FALSE}
betadisper(
  beta_q1n_phylum$S,
  all_samples %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))) %>%
    pull(Species)
) %>%
  permutest(.)

adonis2(
  beta_q1n_phylum$S ~ Species*method,
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))),
  permutations = 999) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q1n_phylum$S ~ Species*method,
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n_phylum$S)) %>% arrange(match(sample,labels(beta_q1n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q1n_phylum$S ~ Species*method,
  by = "terms",
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n_phylum$S)) %>% arrange(match(sample,labels(beta_q1n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()

adonis2(
  beta_q1n_phylum$S ~ Species+method,
  by = "margin",
  data = all_samples %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))),
  permutations = 999,
  strata = all_samples %>% filter(sample %in% labels(beta_q1n_phylum$S)) %>% arrange(match(sample,labels(beta_q1n_phylum$S))) %>% pull(id)) %>%
  broom::tidy() %>%
  tt()
```