# Beta diversity

```{r knitr_options_beta, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(24022025)
```

## Metagenomics
```{r load_data_beta}
load("resources/metagenomics/data_filtered.Rdata")
species_color <- c("#e5bd5b", "#6b7398", "#76b183")
```

***Identifying outliers***
```{r beta_div_nmds_neutral_plot_out, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(
    x = NMDS1,
    y = NMDS2,
    color = Species
  )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")+geom_text_repel(aes(label = sample), size=3)
```

```{r}
load("resources/metagenomics/beta_filtered.Rdata")
sample_metadata <- sample_metadata %>%
  filter(!sample %in% c("H06", "H20"))
genome_counts_filt <- genome_counts_filt %>%
  select(-H06, -H20) %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0))
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)

genome_counts_filt_beta <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  rownames_to_column(., "genome")

genome_tree <- keep.tip(genome_tree, tip = genome_counts_filt_beta$genome)
beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

genome_gifts1 <- genome_gifts[genome_counts_filt_beta$genome, ]
genome_gifts1 <- genome_gifts1[, colSums(genome_gifts1 != 0) > 0]

dist <- genome_gifts1 %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

beta_q1f <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

```{r save_beta, comment="", message=FALSE,echo=FALSE,warning=FALSE, eval=FALSE}
save(beta_q0n, 
     beta_q1n, 
     beta_q1p, 
     beta_q1f, 
     file = "resources/metagenomics/beta_filtered.Rdata")
```

### Permanova

***Richness***
```{r permanova, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q0n$S, sample_metadata$Species) %>% permutest(.)
adonis2(
  beta_q0n$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_neutral_metagenomics, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q0n$S, sample_metadata$Species, perm = 999)
```

***Neutral***
```{r permanova_neutral, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q1n$S, sample_metadata$Species) %>% permutest(.)

adonis2(
  beta_q1n$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_neutral_metagenomics2, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q1n$S, sample_metadata$Species, perm = 999)
```

***Phylogenetic***
```{r permanova_phylo, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q1p$S, sample_metadata$Species) %>% permutest(.)
adonis2(
  beta_q1p$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q1p$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_phylo, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(beta_q1p$S, sample_metadata$Species, perm = 999)
```

***Functional***
```{r permanova_func, comment="", message=FALSE, warning=FALSE}
betadisper(beta_q1f$S, sample_metadata$Species) %>% permutest(.)
adonis2(
  beta_q1f$S ~ Species,
  by = "terms",
  data = sample_metadata %>% arrange(match(sample, labels(beta_q1f$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```

### NMDS

***Richness***
```{r beta_div_nmds_richness_plot, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

***Neutral***
```{r beta_div_nmds_neutral_plot, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(
    x = NMDS1,
    y = NMDS2,
    color = Species
  )) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

#### Phylogenetic diversity

```{r beta_div_nmds_phylo_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1p$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

***Functional diversity***

```{r beta_div_nmds_funct_plot, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1f$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

## Amplicon

```{r div_load_data_ampli2, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```

```{r beta_div_ampli, comment="", message=FALSE, warning=FALSE}
beta_q0n <- genome_counts_filt %>%
#  select(-H06, -H20) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
#  select(-H06, -H20) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

### Permanova

***Richness***
```{r permanova_ampli, comment="", message=FALSE, warning=FALSE}
betadisper(
  beta_q0n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species)
) %>%
  permutest(.)

adonis2(
  beta_q0n$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```
```{r pairwise_rich, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(
  beta_q0n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species),
  perm = 999
)
```

```{r permanova_neutral_ampli, comment="", message=FALSE, warning=FALSE}
#Neutral diversity
betadisper(
  beta_q1n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species)
) %>%
  permutest(.)

adonis2(
  beta_q1n$S ~ Species,
  by = "terms",
  data = sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```

```{r pairwise_neutral, comment="", message=FALSE, warning=FALSE}
pairwise.adonis(
  beta_q1n$S,
  sample_metadata %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species),
  perm = 999
)
```

### NMDS

***Richness***
```{r beta_div_nmds_richness_plot_ampli, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

***Neutral***
```{r beta_div_nmds_neutral_plot_ampli, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  scale_color_manual(values = species_color) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

## Comparison 

```{r beta_div_nmds_richness_plot_ampli_comparison_general, comment="", message=FALSE, warning=FALSE}
sample_metadata_amplicon <- sample_metadata %>% 
  mutate(method="amplicon")

rich_beta_amplicon <- beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method, individual) 

neutral_beta_amplicon <- beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%  
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method,individual)
```
```{r}
load("resources/metagenomics/data_filtered.Rdata")
load("resources/metagenomics/beta_filtered.Rdata")
sample_metadata <- sample_metadata %>%
  filter(!sample %in% c("H06", "H20"))
```

```{r beta_meta_general, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% 
  mutate(method="shotgun")

rich_beta_meta <- beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample))%>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)

neutral_beta_meta <- beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)
```

#### Richness
```{r nmds_rich_general, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_richness_together <- rbind(rich_beta_amplicon, rich_beta_meta) %>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")
beta_richness_together %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species & Method")

```

***Differences between species and methods***
```{r}
metadata_both <- beta_richness_together %>%
  select(sample, Species, method, individual, id)
dist_matrix <- dist(beta_richness_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_richness_together$sample
colnames(dist_matrix) <- beta_richness_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```

#### Neutral
```{r nmds_neu_general, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_neutral_together <- rbind(neutral_beta_amplicon, neutral_beta_meta) %>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")

beta_neutral_together %>%
  filter(sample != "E31_ampli") %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  scale_color_manual(values = beta_colors) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")
```

***Differences between species and methods***
```{r}
metadata_both <- beta_neutral_together %>%
  select(sample, Species, method, individual, id)
dist_matrix <- dist(beta_neutral_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_neutral_together$sample
colnames(dist_matrix) <- beta_neutral_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```





# Different abundance analysis
```{r merge_p, comment="", message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
phyloseq_metagenome_filtered <- subset_samples(phyloseq_metagenome_filtered, !sample_names(phyloseq_metagenome_filtered) %in% c("H06", "H20"))
phyloseq_metagenome_filtered <- prune_taxa(taxa_sums(phyloseq_metagenome_filtered)>0, phyloseq_metagenome_filtered)
sample_data(phyloseq_metagenome_filtered)$method <- "shotgun"
sample_data(phyloseq_metagenome_filtered)$sample_ID <- rownames(sample_data(phyloseq_metagenome_filtered))

# otu_table(phyloseq_metagenome_filtered) <- otu_table(
#   apply(otu_table(phyloseq_metagenome_filtered), 2, as.integer), 
#   taxa_are_rows = taxa_are_rows(phyloseq_metagenome_filtered)  # Preserve original orientation
# )

phyloseq_metagenome_filtered_phylum <- tax_glom(phyloseq_metagenome_filtered, "phylum")

load("resources/amplicon/data_standard_filt.Rdata")
sample_data(phyloseq_16S_stand)$sample_ID <- rownames(sample_data(phyloseq_16S_stand))
sample_names(phyloseq_16S_stand) <- paste0(sample_names(phyloseq_16S_stand), "_ampli")
sample_data(phyloseq_16S_stand)$method <- "amplicon"

phyloseq_16S_stand_phylum <- tax_glom(phyloseq_16S_stand, "phylum")

merge_phylo <- merge_phyloseq(phyloseq_16S_stand_phylum,phyloseq_metagenome_filtered_phylum)

```

```{r merge_otu, comment="", message=FALSE, warning=FALSE}
otu1 <- as.data.frame(otu_table(phyloseq_metagenome_filtered_phylum)) %>%
  mutate_all(as.integer)
tax1 <- as.data.frame(tax_table(phyloseq_metagenome_filtered_phylum)) %>% 
  select(phylum)
otu1$phylum <- tax1$phylum

otu2 <- as.data.frame(otu_table(phyloseq_16S_stand_phylum))
tax2 <- as.data.frame(tax_table(phyloseq_16S_stand_phylum)) %>% select(phylum)
otu2$phylum <- tax2$phylum

merged_otu <- full_join(otu1, otu2, by = "phylum") %>%
  mutate(across(everything(), ~replace_na(., 0)))%>%
  group_by(phylum) %>%
  summarise(across(everything(), sum, na.rm = TRUE))
merged_otu_mat <- as.matrix(merged_otu %>% select(-phylum))
rownames(merged_otu_mat) <- merged_otu$phylum
otu_table_merged <- otu_table(merged_otu_mat, taxa_are_rows = TRUE)

merged_physeq_phylum <- phyloseq(otu_table_merged, sample_data(merge_phylo))
merged_physeq_phylum <- prune_taxa(taxa_sums(merged_physeq_phylum)>0, merged_physeq_phylum)

```

```{r}
ancom_rand_output_struc = ancombc2(data = merged_physeq_phylum, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  rand_formula = "(1|sample_ID)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total number of structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc$zero_ind
structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")
structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]
structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("Amplicon"=1) %>% 
  rename("Shotgun"=2)
nrow(structural_zero_groups)
```

***Structural zeros in Amplicon***
```{r}
structural_zero_groups %>% 
  filter(Amplicon==TRUE) %>% 
  nrow() %>% 
  cat()

structural_zero_groups %>% 
  filter(Amplicon==TRUE)
```

***Structural zeros in Shotgun***
```{r}
structural_zero_groups %>% 
  filter(Shotgun==TRUE) %>% 
  nrow() %>% 
  cat()

structural_zero_groups %>% 
  filter(Shotgun==TRUE)
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc$res %>% 
  filter(p_methodshotgun<0.05)
```

***Phyla not different***
```{r}
ancom_rand_output_struc$res %>% 
  filter(p_methodshotgun>0.05)
```

