# Differential abundance

```{r}
load("resources/amplicon/data_standard_filt.Rdata")

sample_metadata_amplicon <- sample_metadata[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding_standard")

genome_metadata_ampli <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

```{r }
load("resources/metagenomics/data_filtered.Rdata")

genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

Merge the table
```{r}
merged_table <- full_join(family_level_agg_ampli,family_level_agg, 
          by = c("family"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

merged_table_phylum <- full_join(phylum_level_agg_ampli,phylum_level_agg, 
          by = c("phylum"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


sample_shot <- sample_metadata %>% 
  select(-c(Habitat,Longitude, Latitude, Region))%>%
  mutate(method="Metagenomics")

all_samples <- rbind(sample_shot, sample_metadata_amplicon)

```

## Phylum level
### Metagenomics

```{r zero_phylum_meta, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_taxonomy <- genome_metadata %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_phylum_metagen <- phyloseq(phylo_genome, phylo_samples, phylo_taxonomy)
```

```{r ancom_rand_phylum_metag, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_phylum_metagen = ancombc2(
  data = physeq_genome_phylum_metagen,
  assay_name = "counts",
  tax_level = "phylum",
  fix_formula = "Species",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "Species",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 20,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)
```

#### Structural zeros

***Total number of structural zeros***
```{r}
# Extract structural zeros
structural_zeros <- ancom_rand_output_phylum_metagen$zero_ind

# Convert to a data frame for better readability
structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

# Identify in which groups each taxon is a structural zero
structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("Eb"=1) %>% 
  rename("Ha"=2) %>% 
  rename("Pk"=3) 

# Total number of structural zeros
nrow(structural_zero_groups)

```

***Structural zeros in EB***
```{r}
structural_zero_groups %>% 
  filter(Eb==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(Eb==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in Ha***
```{r}
structural_zero_groups %>% 
  filter(Ha==TRUE) %>% 
  rownames_to_column(., "phylum")%>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in Pk***
```{r}
structural_zero_groups %>% 
  filter(Pk==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  select(phylum) %>% 
  pull()
```

#### Enriched data
```{r labeling_phyl_metag, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
res_pair_phylum <- ancom_rand_output_phylum_metagen$res_pair
# Get log-fold change and q-value columns
lfc_cols <- grep("^lfc_", colnames(res_pair_phylum), value = TRUE)
q_cols   <- gsub("^lfc_", "q_", lfc_cols)

# Build long table by pairing lfc and q for each contrast
plot_data_phylum <- lfc_cols %>%
  setNames(nm = .) %>%
  purrr::imap_dfr(function(lfc_col, name) {
    q_col <- gsub("^lfc_", "q_", lfc_col)
    
    res_pair_phylum %>%
      select(taxon, !!lfc_col, !!q_col) %>%
      rename(
        lfc = !!lfc_col,
        q   = !!q_col
      ) %>%
      mutate(contrast_key = name)
  }) %>%
  filter(!is.na(lfc), !is.na(q), q < 0.05) %>%
  mutate(
    # Manually re-label contrasts for clarity
    contrast = case_when(
      contrast_key == "lfc_SpeciesHa" ~ "H. ariel vs C. bottae",
      contrast_key == "lfc_SpeciesPk" ~ "P. kuhlii vs C. bottae",
      contrast_key == "lfc_SpeciesPk_SpeciesHa" ~ "P. kuhlii vs H. ariel",
      TRUE ~ contrast_key
    ),
    direction = ifelse(lfc > 0, paste("Enriched in", sub(" vs.*", "", contrast)),
                             paste("Enriched in", sub(".*vs ", "", contrast)))
  )
plot_data_phylum
```

### Metabarcoding
```{r div_load_data_ampli2_metab, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```
```{r zero_phylo1_metab, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata_amplicon %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli"))%>%
  rownames_to_column(., "genome") %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_taxonomy <- genome_metadata_ampli %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_phylum_metab <- phyloseq(phylo_genome, phylo_samples, phylo_taxonomy)
```

```{r ancom_rand_phylum_metab, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_phylum_metab = ancombc2(
  data = physeq_genome_phylum_metab,
  assay_name = "counts",
  tax_level = "phylum",
  fix_formula = "Species",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "Species",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 20,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)
```

#### Structural zeros

***Total number of structural zeros***
```{r}
# Extract structural zeros
structural_zeros <- ancom_rand_output_phylum_metab$zero_ind

# Convert to a data frame for better readability
structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

# Identify in which groups each taxon is a structural zero
structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("Eb"=1) %>% 
  rename("Ha"=2) %>% 
  rename("Pk"=3) 

# Total number of structural zeros
nrow(structural_zero_groups)
```

#### Enriched data
```{r  labeling_phyl_metab, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
res_pair_phylum <- ancom_rand_output_phylum_metab$res_pair
# Get log-fold change and q-value columns
lfc_cols <- grep("^lfc_", colnames(res_pair_phylum), value = TRUE)
q_cols   <- gsub("^lfc_", "q_", lfc_cols)

# Build long table by pairing lfc and q for each contrast
plot_data_phylum <- lfc_cols %>%
  setNames(nm = .) %>%
  purrr::imap_dfr(function(lfc_col, name) {
    q_col <- gsub("^lfc_", "q_", lfc_col)
    
    res_pair_phylum %>%
      select(taxon, !!lfc_col, !!q_col) %>%
      rename(
        lfc = !!lfc_col,
        q   = !!q_col
      ) %>%
      mutate(contrast_key = name)
  }) %>%
  filter(!is.na(lfc), !is.na(q), q < 0.05) %>%
  mutate(
    # Manually re-label contrasts for clarity
    contrast = case_when(
      contrast_key == "lfc_SpeciesHa" ~ "H. ariel vs C. bottae",
      contrast_key == "lfc_SpeciesPk" ~ "P. kuhlii vs C. bottae",
      contrast_key == "lfc_SpeciesPk_SpeciesHa" ~ "P. kuhlii vs H. ariel",
      TRUE ~ contrast_key
    ),
    direction = ifelse(lfc > 0, paste("Enriched in", sub(" vs.*", "", contrast)),
                             paste("Enriched in", sub(".*vs ", "", contrast)))
  )
plot_data_phylum
```

### Differences between methods accross species

```{r}
phylo_genome <- merged_table_phylum %>%
  column_to_rownames("phylum") %>% 
  otu_table(., taxa_are_rows = TRUE)

phylo_samples <- all_samples%>% 
  mutate(individual=sample)%>%
  column_to_rownames("sample") %>% 
  sample_data()

merged_physeq_phylum <- phyloseq(phylo_genome, phylo_samples)
```

#### Eptesicus
```{r}
physeq_eb <- subset_samples(merged_physeq_phylum, Species == "Eb")
physeq_eb <- prune_taxa(taxa_sums(physeq_eb)>0, physeq_eb)
```

```{r}
ancom_rand_output_struc_eb = ancombc2(data = physeq_eb, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```


***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_eb$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)

```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_eb$res %>% 
  filter(q_methodMetagenomics<0.05)
```

#### Pipistrellus
```{r}
physeq_pk <- subset_samples(merged_physeq_phylum, Species == "Pk")
physeq_pk <- prune_taxa(taxa_sums(physeq_pk)>0, physeq_pk)
```

```{r}
ancom_rand_output_struc_pk = ancombc2(data = physeq_pk, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_pk$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)
```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_pk$res %>% 
  filter(q_methodMetagenomics<0.05)
```

#### Hypsugo
```{r}
physeq_ha <- subset_samples(merged_physeq_phylum, Species == "Ha")
physeq_ha <- prune_taxa(taxa_sums(physeq_ha)>0, physeq_ha)
```

```{r}
ancom_rand_output_struc_ha = ancombc2(data = physeq_ha, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_ha$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)

```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_ha$res %>% 
  filter(q_methodMetagenomics<0.05)
```