# Amplicon t3_p10 filtering vs genome resolved metagenomics

```{r knitr_options_comp, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(24022025)
```

Check filtered ASVs
```{r}
load("resources/amplicon/selected_filtering.Rdata")
load("resources/amplicon/data_standard_filt.Rdata")

filtered <- genome_counts_filt %>% 
  filter(!genome %in% genome_counts_filt_amplicon$genome)

filtered %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>% 
  rownames_to_column(., "genome") %>% 
  left_join(genome_metadata, by="genome") %>% 
  group_by(phylum)

```

```{r load_data_comp}
load("resources/amplicon/selected_filtering.Rdata")
load("resources/metagenomics/data_filtered.Rdata")
```

## Taxonomy

Stacked bar chart with filtered amplicon data
```{r taxonomy_barplot_ampli_filt, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt_amplicon %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata_amplicon, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata_amplicon, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Eptesicus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 6),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 2))
```

***Number of MAGs in metagenomic data***
```{r}
cat(nrow(genome_metadata))
```

***Number of ASVs in amplicon data***
```{r}
cat(nrow(genome_metadata_amplicon))
```

### Differences at phylum level
***Number of phyla in metagenomic data***
```{r}
cat(genome_metadata %>% 
  distinct(domain,phylum) %>% 
  count() %>% 
    pull())
```

***Number of phyla in amplicon data***
```{r}
cat(genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  count()%>% 
    pull())
```

***Shared phyla***
```{r}
meta_phylum <- genome_metadata %>% 
  distinct(domain,phylum)

genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(phylum %in% meta_phylum$phylum)
```

***Only in metagenomic data***
```{r}
ampli_phylum <- genome_metadata_amplicon %>% 
  distinct(domain,phylum)
genome_metadata %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% genome_metadata_amplicon$phylum)
```

***Only in amplicon data***
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% meta_phylum$phylum)
```

### Differences at family level
***Number of families in metagenomic data***
```{r}
cat(genome_metadata %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

***Number of families in amplicon data***
```{r}
cat(genome_metadata_amplicon %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

***Shared families***
```{r}
meta_family <- genome_metadata %>% 
  distinct(domain,phylum, family)
genome_metadata_amplicon %>% 
  distinct(domain,phylum, family) %>% 
  filter(family %in% meta_family$family)
```

***Only in metagenomic data***
```{r}
ampli_family <- genome_metadata_amplicon %>% 
  distinct(domain,phylum,family)
genome_metadata %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% ampli_family$family)
```

***Only in amplicon data***
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% meta_family$family)
```

### Differences in top ASVs and MAGs
Top 20 MAGs
```{r}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  mutate(total = rowSums(across(-genome))) %>%
  arrange(desc(total)) %>%
  slice_head(n = 20) %>% 
  select(genome, total) %>% 
  left_join(genome_metadata[1:8])
```

Top 20 ASVs
```{r}
genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  mutate(total = rowSums(across(-genome))) %>%
  arrange(desc(total)) %>%
  slice_head(n = 20) %>% 
  select(genome, total) %>% 
  left_join(genome_metadata_amplicon)
```

Phylum level
```{r}
topMAG <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum) %>% 
  arrange(desc(total_mag))%>%
  slice_head(n = 10)
```

```{r}
topASV <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 10)

table <- full_join(topMAG,topASV, by="phylum") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
  select(-total_asv, -total_mag) %>% 
  pivot_longer(cols = c(MAG, ASV),
               names_to = "source",
               values_to = "abundance")
```

```{r figure_bar_phylum_filt, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
ggplot(table, aes(x = reorder(phylum, -abundance), y = abundance, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Phylum", y = "Total abundance", fill = "Source") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18))
```
```{r figure_bubble_phylum_filt, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
ggplot(table, aes(x = source, y = phylum, size = abundance, color = phylum)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(
    name = "Abundance",
    range = c(0.01, 30),   # 0 ensures no bubble at abundance = 0
    breaks = seq(0, 35, by = 5),   # Controls legend labels (your 0–5–10–...–30)
    limits = c(0.01, NA)   # keep full range of values
  )  +
    scale_color_manual(values = phylum_colors) +
    coord_flip()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))+
  labs(x = "Source", y = "Phylum",
    title = "Metabarcoding with selected threshold")+
  guides(color = "none", size = guide_legend(
      override.aes = list(size = c(2, 4, 6, 8, 10, 12)) # smaller bubbles in legend
    )
  )
```
```{r figure_scat_phylum_filt, fig.height=5, fig.width=7, fig.fullwidth=TRUE}
full_join(topMAG,topASV, by="phylum") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
  ggplot(., aes(x = MAG, y = ASV, color = phylum, label = phylum)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
  labs(
    x = "MAG abundance",
    y = "ASV abundance",
    color = "Phylum",
    title = "Comparison of MAG and ASV abundances by phylum"
  ) +
  theme_minimal()
```

Family level
```{r}
topFMAG <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum, family) %>% 
  arrange(desc(total_mag)) %>%
  slice_head(n = 20)
```

```{r}
topFASV <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum, family) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 20)
```
```{r}
table_family <- full_join(topFMAG,topFASV, by="family") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
    rename(
    phylum_mag = phylum.x,
   phylum_asv =phylum.y,
  )%>%
  mutate(phylum_mag = coalesce(phylum_mag, phylum_asv))%>%
  pivot_longer(cols = c(MAG, ASV),
               names_to = "source",
               values_to = "abundance")
```

```{r figure_bar_phylum_filt2, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
ggplot(table_family, aes(x = reorder(family, -abundance), y = abundance, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "family", y = "Total abundance", fill = "Source") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18))
```
```{r figure_bubble_phylum_filt2, fig.height=7, fig.width=20, fig.fullwidth=TRUE}
ggplot(table_family, aes(x = source, y = family, size = abundance, color = phylum_mag)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(
    name = "Abundance",
    range = c(0.01, 20),   # 0 ensures no bubble at abundance = 0
    breaks = seq(0, 20, by = 2),   # Controls legend labels (your 0–5–10–...–30)
    limits = c(0.01, NA)   # keep full range of values
  )  +
    scale_color_manual(values = phylum_colors) +
    coord_flip()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))+
  labs(x = "Source", y = "Phylum",
    title = "Metabarcoding with standard filtering")+
  guides(color = "none", size = guide_legend(
      override.aes = list(size = c(2, 4, 6, 8, 10)) # smaller bubbles in legend
    )
  )
```

```{r figure_scat_phylum_filt2, fig.height=8, fig.width=12, fig.fullwidth=TRUE}
full_join(topFMAG,topFASV, by="family") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
    mutate(phylum_mag = coalesce(phylum.x, phylum.y))%>%
  ggplot(., aes(x = MAG, y = ASV, color = phylum_mag, label = family)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
      scale_color_manual(values = phylum_colors) +
  labs(
    x = "MAG abundance",
    y = "ASV abundance",
    color = "Phylum",
    title = "Metabarcoding with selected filtering"
  ) +
  theme_minimal()
```

Genus level
```{r}
top20MAG <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family, genus) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum, family, genus) %>% 
  arrange(desc(total_mag)) %>%
  slice_head(n = 20)
```

```{r}
top20ASV <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum, family, genus) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum, family, genus) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 20)
```

```{r}
table <- full_join(top20MAG,top20ASV, by="genus") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
    rename(
    phylum_mag = phylum.x,
   family_mag= family.x,
   phylum_asv =phylum.y,
  family_asv=family.y,
  )%>%
  pivot_longer(cols = c(MAG, ASV),
               names_to = "source",
               values_to = "abundance")

ggplot(table, aes(x = reorder(genus, -abundance), y = abundance, fill = source)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Genus", y = "Total abundance", fill = "Source") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(table, aes(x = source, y = genus, size = abundance, color = phylum_mag)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(name = "Abundance") +
  labs(x = "Source (MAG / ASV)", y = "Genus", color = "Phylum") +
  theme_minimal()
  
```

## Alpha Diversity

### Merge both data

```{r alpha_cal_filt, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

amplicon_alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>% 
  mutate(method="amplicon")
```

```{r load_data_alpha_meta_filt}
metagenome_alpha_div <- readRDS("resources/metagenomics/metagenome_alpha_div.rds") %>% 
  select(sample, richness, neutral, method)

join_alpha <- rbind(metagenome_alpha_div,amplicon_alpha_div) %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

### Comparison

#### Differences between methods and species

***Richness***
```{r rich_MM_comp_species_filt, comment="", message=FALSE, warning=FALSE}
Model_richness <- lm(formula = richness ~ Species*method, data = join_alpha)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

***Neutral***
```{r neutral_MM_com_species_filt, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species*method, data = join_alpha)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)

```

##### Metagenomic data

```{r meta2, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- metagenome_alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

***Richness***
```{r rich_MM_species2, comment="", message=FALSE, warning=FALSE}
Model_richness <- lm(formula = richness ~ Species, data = alpha_div_meta)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

***Neutral***
```{r neutral_MM2, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_meta)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

##### Amplicon data

```{r meta_ampli_filt, comment="", message=FALSE, warning=FALSE}
alpha_div_ampli <- amplicon_alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

***Richness***
```{r rich_MM_ampli_species_filt, comment="", message=FALSE, warning=FALSE}
Model_richness <- lm(formula = richness ~ Species, data = alpha_div_ampli)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

***Neutral***
```{r neutral_MM_ampli_species_filt, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_ampli)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

## Beta diversity

### Prepare the data
```{r}
load("resources/amplicon/selected_filtering.Rdata")
```

***Calculate beta diversity in amplicon data***
```{r beta_div_ampli_comparison, comment="", message=FALSE, warning=FALSE}
beta_q0n_ampli <- genome_counts_filt_amplicon %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n_ampli <- genome_counts_filt_amplicon %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

***Check for outliers in amplicon data***
```{r}
beta_q1n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")+geom_text_repel(aes(label = sample), size=3)

```

***E31 has been considered an outlier, so it has been removed, and the diversity has been calculated again***
```{r beta_div_ampli_new_comparison, comment="", message=FALSE, warning=FALSE}
beta_q0n_ampli <- genome_counts_filt_amplicon %>%
    select(-E31) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n_ampli <- genome_counts_filt_amplicon %>%
    select(-E31) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

```{r beta_div_nmds_richness_plot_ampli_comparison, comment="", message=FALSE, warning=FALSE}
sample_metadata_amplicon <- sample_metadata_amplicon %>% 
  mutate(method="amplicon")

rich_beta_amplicon <- beta_q0n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method, individual) 


neutral_beta_amplicon <- beta_q1n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%  
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method,individual)
```

***Calculate beta diversity in metagenomic data***
```{r}
load("resources/metagenomics/data_filtered.Rdata")
load("resources/metagenomics/beta_filtered.Rdata")
sample_metadata <- sample_metadata %>%
  filter(!sample %in% c("H06", "H20"))
```

```{r beta_meta, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% 
  mutate(method="shotgun")

rich_beta_meta <- beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample))%>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)


neutral_beta_meta <- beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)
```

### Comparison

#### Richness
```{r nmds_rich, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_richness_together <- rbind(rich_beta_amplicon,rich_beta_meta)%>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")
beta_richness_together %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species & Method")

```

***Differences between species and methods***
```{r}
metadata_both <- beta_richness_together %>% 
  select(sample, Species, method,individual, id)
dist_matrix <- dist(beta_richness_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_richness_together$sample
colnames(dist_matrix) <- beta_richness_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```

#### Neutral
```{r nmds_neu, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_neutral_together <- rbind(neutral_beta_amplicon,neutral_beta_meta) %>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")

beta_neutral_together %>%
  filter(sample!="E31_ampli") %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  scale_color_manual(values = beta_colors)+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")

```

***Differences between species and methods***
```{r}
metadata_both <- beta_neutral_together %>% 
  select(sample, Species, method,individual, id)
dist_matrix <- dist(beta_neutral_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_neutral_together$sample
colnames(dist_matrix) <- beta_neutral_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```

## Differential abundace analysis at phylum level

```{r}
phylo_samples <- sample_metadata %>% 
  mutate(individual=sample)%>%
  column_to_rownames("sample") %>% 
  mutate(method="shotgun") %>%
  select(1,8,9) %>%
  sample_data()
phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_filtered_phylum <- tax_glom(physeq_genome_filtered, "phylum")
```


```{r}
phylo_samples <- sample_metadata_amplicon %>% 
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli"))%>%
  column_to_rownames("sample")  %>% 
  mutate(method="amplicon")%>% 
  select(12,15,16) %>%
  sample_data()
phylo_genome <- genome_counts_filt_amplicon%>%
  rename_with(~ paste0(., "_ampli")) %>%
  select(one_of(c("genome_ampli",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome_ampli, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome_ampli") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata_amplicon %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered_ampl <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_filtered_ampl_phylum <- tax_glom(physeq_genome_filtered_ampl, "phylum")

merge_phylo <- merge_phyloseq(physeq_genome_filtered_ampl_phylum,physeq_genome_filtered_phylum)
```

```{r}
#create the merge phyloseq object at phylum level
otu1 <- as.data.frame(otu_table(physeq_genome_filtered_phylum))
tax1 <- as.data.frame(tax_table(physeq_genome_filtered_phylum)) %>% select(phylum)
otu1$phylum <- tax1$phylum

otu2 <- as.data.frame(otu_table(physeq_genome_filtered_ampl_phylum))
tax2 <- as.data.frame(tax_table(physeq_genome_filtered_ampl_phylum)) %>% select(phylum)
otu2$phylum <- tax2$phylum

merged_otu <- full_join(otu1, otu2, by = "phylum") %>%
  mutate(across(everything(), ~replace_na(., 0)))%>%
  group_by(phylum) %>%
  summarise(across(everything(), sum, na.rm = TRUE))
merged_otu_mat <- as.matrix(merged_otu %>% select(-phylum))
rownames(merged_otu_mat) <- merged_otu$phylum
otu_table_merged <- otu_table(merged_otu_mat, taxa_are_rows = TRUE)

merged_physeq_phylum <- phyloseq(otu_table_merged, sample_data(merge_phylo))

```

### Eptesicus
```{r}
physeq_eb <- subset_samples(merged_physeq_phylum, Species == "Eb")
physeq_eb <- prune_taxa(taxa_sums(physeq_eb)>0, physeq_eb)
```

```{r}
ancom_rand_output_struc_eb = ancombc2(data = physeq_eb, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_eb$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)

```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_eb$res %>% 
  filter(p_methodshotgun<0.05)
```

### Pipistrellus
```{r}
physeq_pk <- subset_samples(merged_physeq_phylum, Species == "Pk")
physeq_pk <- prune_taxa(taxa_sums(physeq_pk)>0, physeq_pk)
```

```{r}
ancom_rand_output_struc_pk = ancombc2(data = physeq_pk, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_pk$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)

```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_pk$res %>% 
  filter(p_methodshotgun<0.05)
```


### Hypsugo (not working random effect)
```{r}
physeq_ha <- subset_samples(merged_physeq_phylum, Species == "Ha")
physeq_ha <- prune_taxa(taxa_sums(physeq_ha)>0, physeq_ha)
```

```{r}
ancom_rand_output_struc_ha = ancombc2(data = physeq_ha, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
#                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
 #                 lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_ha$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("shotgun"=2)

nrow(structural_zero_groups)

```

***Structural zeros in amplicon***
```{r}

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in shotgun***
```{r}
structural_zero_groups %>% 
  filter(shotgun==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_ha$res %>% 
  filter(p_methodshotgun<0.05)
```
