# Amplicon t3_p10 filtering vs genome resolved metagenomics

```{r knitr_options_comp, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(24022025)
```

```{r load_data_comp}
load("resources/amplicon/selected_filtering.Rdata")
load("resources/metagenomics/data_filtered.Rdata")
```

## Taxonomy

***Number of MAGs in metagenomic data***
```{r}
cat(nrow(genome_metadata))
```

***Number of ASVs in amplicon data***
```{r}
cat(nrow(genome_metadata_amplicon))
```


### Phylum
***Number of phyla in metagenomic data***
```{r}
cat(genome_metadata %>% 
  distinct(domain,phylum) %>% 
  count() %>% 
    pull())
```

***Number of phyla in amplicon data***
```{r}
cat(genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  count()%>% 
    pull())
```

***Shared phyla***
```{r}
meta_phylum <- genome_metadata %>% 
  distinct(domain,phylum)

genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(phylum %in% meta_phylum$phylum)
```

***Only in metagenomic data***
```{r}
ampli_phylum <- genome_metadata_amplicon %>% 
  distinct(domain,phylum)
genome_metadata %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% genome_metadata_amplicon$phylum)
```

***Only in amplicon data***
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% meta_phylum$phylum)
```

### Family
***Number of families in metagenomic data***
```{r}
cat(genome_metadata %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

***Number of families in amplicon data***
```{r}
cat(genome_metadata_amplicon %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

***Shared families***
```{r}
meta_family <- genome_metadata %>% 
  distinct(domain,phylum, family)
genome_metadata_amplicon %>% 
  distinct(domain,phylum, family) %>% 
  filter(family %in% meta_family$family)
```

***Only in metagenomic data***
```{r}
ampli_family <- genome_metadata_amplicon %>% 
  distinct(domain,phylum,family)
genome_metadata %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% ampli_family$family)
```

***Only in amplicon data***
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% meta_family$family)
```

## Alpha Diversity

### Calculate for the selected amplicon data

```{r alpha_cal, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

amplicon_alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>% 
  mutate(method="amplicon")
```

```{r load_data_alpha_meta}
metagenome_alpha_div <- readRDS("resources/metagenomics/metagenome_alpha_div.rds") %>% 
  select(sample, richness, neutral, method)

join_alpha <- rbind(metagenome_alpha_div,amplicon_alpha_div) %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

### Species differences
#### Metagenomic data

```{r meta, comment="", message=FALSE, warning=FALSE}
alpha_div_meta <- metagenome_alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

***Richness***
```{r rich_MM_species, comment="", message=FALSE, warning=FALSE}
Model_richness <- lm(formula = richness ~ Species, data = alpha_div_meta)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

***Neutral***
```{r neutral_MM, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_meta)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

### Amplicon data

```{r meta_ampli, comment="", message=FALSE, warning=FALSE}
alpha_div_ampli <- amplicon_alpha_div %>%
  left_join(sample_metadata, by = join_by(sample == sample))
```

***Richness***
```{r rich_MM_ampli_species, comment="", message=FALSE, warning=FALSE}
Model_richness <- lm(formula = richness ~ Species, data = alpha_div_ampli)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

***Neutral***
```{r neutral_MM_ampli_species, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_ampli)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

### Differences between methods in each species
```{r}
join_alpha %>%
  group_by(sample) %>%
  filter(all(c("shotgun", "amplicon") %in% method))
```

#### Richness
***Eptesicus***
```{r}
join_richness_eb <- join_alpha %>% 
  filter(Species=="Eb") %>% 
  select(sample,method,richness) 


ggboxplot(join_richness_eb, x = "method", y = "richness", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "richness", xlab = "method")
```
```{r}
shapiro.test(join_richness_eb$richness)

eb_result <- join_richness_eb %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = richness) %>%
  drop_na() %>%  # Ensure no missing pairs
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(eb_result$wilcox_result[[1]])
```

***Hypsugo***
```{r}
join_richness_ha <- join_alpha %>% 
  filter(Species=="Ha") %>% 
  select(sample,method,richness)

ggboxplot(join_richness_ha, x = "method", y = "richness", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "richness", xlab = "method")
```
```{r}
shapiro.test(join_richness_ha$richness)

ha_result <- join_richness_ha %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = richness) %>%
  drop_na() %>%  # Ensure no missing pairs
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(ha_result$wilcox_result[[1]])
```

***Pipistrellus***
```{r}
join_richness_pk <- join_alpha %>% 
  filter(Species=="Pk") %>% 
  select(sample,method,richness)

ggboxplot(join_richness_pk, x = "method", y = "richness", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "richness", xlab = "method")
```
```{r}
shapiro.test(join_richness_pk$richness)

pk_result <- join_richness_pk %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = richness) %>%
  drop_na() %>%  # Ensure no missing pairs
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(pk_result$wilcox_result[[1]])
```

#### Neutral
***Eptesicus***
```{r}
join_neutral_eb <- join_alpha %>% 
  filter(Species=="Eb") %>% 
  select(sample,method,neutral)

ggboxplot(join_neutral_eb, x = "method", y = "neutral", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "neutral", xlab = "method")
```
```{r}
shapiro.test(join_neutral_eb$neutral)

eb_result <- join_neutral_eb %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = neutral) %>%
  drop_na() %>%  # Ensure no missing pairs
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(eb_result$wilcox_result[[1]])
```

***Hypsugo***
```{r}
join_neutral_ha <- join_alpha %>% 
  filter(Species=="Ha") %>% 
  select(sample,method,neutral)

ggboxplot(join_neutral_ha, x = "method", y = "neutral", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "neutral", xlab = "method")
```
```{r}
shapiro.test(join_neutral_ha$neutral)

ha_result <- join_neutral_ha %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = neutral) %>%
  drop_na() %>%
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(ha_result$wilcox_result[[1]])
```

***Pipistrellus***
```{r}
join_neutral_pk <- join_alpha %>% 
  filter(Species=="Pk") %>% 
  select(sample,method,neutral)

ggboxplot(join_neutral_pk, x = "method", y = "neutral", 
          color = "method", palette = c("#00AFBB", "#E7B800"),
          order = c("amplicon", "shotgun"),
          ylab = "neutral", xlab = "method")
```
```{r}
shapiro.test(join_neutral_pk$neutral)

pk_result <- join_neutral_pk %>%
  filter(method %in% c("shotgun", "amplicon")) %>%
  pivot_wider(names_from = method, values_from = neutral) %>%
  drop_na() %>%
  summarize(wilcox_result = list(wilcox.test(shotgun, amplicon, paired = TRUE, exact = FALSE)))

print(pk_result$wilcox_result[[1]])
```

## Beta diversity

### Amplicon
```{r}
load("resources/amplicon/selected_filtering.Rdata")
```

```{r beta_div_ampli_comparison, comment="", message=FALSE, warning=FALSE}
beta_q0n_ampli <- genome_counts_filt_amplicon %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n_ampli <- genome_counts_filt_amplicon %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

***Check for outliers in amplicon data***
```{r}
beta_q1n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species)) +
  geom_point(size = 4) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")+geom_text_repel(aes(label = sample), size=3)

```

***E31 has been considered an outlier, so it has been removed***
```{r beta_div_ampli_new_comparison, comment="", message=FALSE, warning=FALSE}
beta_q0n_ampli <- genome_counts_filt_amplicon %>%
    select(-E31) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n_ampli <- genome_counts_filt_amplicon %>%
    select(-E31) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  hillpair(., q = 1)
```

```{r beta_div_nmds_richness_plot_ampli_comparison, comment="", message=FALSE, warning=FALSE}
sample_metadata_amplicon <- sample_metadata_amplicon %>% 
  mutate(method="amplicon")

rich_beta_amplicon <- beta_q0n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method, individual) 


neutral_beta_amplicon <- beta_q1n_ampli$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%  
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli")) %>% 
  select(sample,NMDS1,NMDS2,Species,method,individual)
```

### Metagenomics
```{r}
load("resources/metagenomics/data_filtered.Rdata")
load("resources/metagenomics/beta_filtered.Rdata")
sample_metadata <- sample_metadata %>%
  filter(!sample %in% c("H06", "H20"))
```

```{r beta_meta, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>% 
  mutate(method="shotgun")

rich_beta_meta <- beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample))%>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)


neutral_beta_meta <- beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  mutate(individual=sample)%>% 
  select(sample,NMDS1,NMDS2,Species,method, individual)
```

### Comparison

#### Richness
```{r nmds_rich, fig.height=10, fig.width=16, fig.fullwidth=TRUE}
beta_richness_together <- rbind(rich_beta_amplicon,rich_beta_meta)%>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")
beta_richness_together %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species & Method")

```

***Differences between species and methods***
```{r}
metadata_both <- beta_richness_together %>% 
  select(sample, Species, method,individual, id)
dist_matrix <- dist(beta_richness_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_richness_together$sample
colnames(dist_matrix) <- beta_richness_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```


#### Neutral
```{r nmds_neu, fig.height=10, fig.width=16, fig.fullwidth=TRUE}
beta_neutral_together <- rbind(neutral_beta_amplicon,neutral_beta_meta) %>%
  mutate(id = paste(Species, method, sep = "_"))
beta_colors <- c("#e5bd5b","#d57d2c","#6b7398","#ea68c3","#76b183","#3c634e")

beta_neutral_together %>%
  filter(sample!="E31_ampli") %>%
  group_by(id) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = id)) +
  geom_point(size = 4) +
  scale_color_manual(values = beta_colors)+
  geom_segment(
    aes(
      x = x_cen,
      y = y_cen,
      xend = NMDS1,
      yend = NMDS2
    ),
    alpha = 0.9,
    show.legend = FALSE
  ) +
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Species")

```

***Differences between species and methods***
```{r}
metadata_both <- beta_neutral_together %>% 
  select(sample, Species, method,individual, id)
dist_matrix <- dist(beta_neutral_together %>% select(NMDS1, NMDS2), method = "euclidean")
dist_matrix <- as.matrix(dist_matrix)
rownames(dist_matrix) <- beta_neutral_together$sample
colnames(dist_matrix) <- beta_neutral_together$sample

adonis2(dist_matrix ~ method*Species,
        data = metadata_both %>% arrange(match(sample,labels(dist_matrix))),
        permutations = 999, by="terms") %>%
        broom::tidy() %>%
        tt()
```


## Differential abundace analysis
### Phylum
```{r}
phylo_samples <- sample_metadata %>% 
  mutate(individual=sample)%>%
  column_to_rownames("sample") %>% 
  select(1,8,9) %>%
  mutate(method="shotgun") %>% 
  sample_data()
phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_filtered_phylum <- tax_glom(physeq_genome_filtered, "phylum")
```
```{r load_data_raw_ampli, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/data_nocopyfilt.Rdata")

genome_counts_selected <- genome_counts_filt %>%
    select(all_of(colnames(genome_counts_filt_amplicon))) %>% 
  filter(genome %in% genome_counts_filt_amplicon$genome) %>%   
  arrange(genome) %>% 
  column_to_rownames("genome")

table <- genome_counts_filt_amplicon %>%
  select(genome, all_of(colnames(genome_counts_selected))) %>%
  mutate(across(all_of(colnames(genome_counts_selected)), ~ ifelse(. > 0, 1, 0)))%>%   
  arrange(genome)%>% 
  column_to_rownames("genome")


genome_counts_amplicon_selected <- table * genome_counts_selected
  

```

```{r}
phylo_samples <- sample_metadata_amplicon %>% 
  mutate(individual=sample)%>%
  mutate(sample = paste0(sample, "_ampli"))%>%
  column_to_rownames("sample")  %>% 
  select(12,15,16) %>%
  mutate(method="amplicon")%>% 
  sample_data()
phylo_genome <- genome_counts_amplicon_selected %>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column("genome")%>% 
  select(one_of(c("genome",rownames(phylo_samples)))) %>%
  filter(rowSums(across(where(is.numeric), ~ . != 0)) > 0) %>%
  select(genome, where(~ is.numeric(.) && sum(.) > 0)) %>%
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata_amplicon %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_filtered_ampl <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_filtered_ampl_phylum <- tax_glom(physeq_genome_filtered_ampl, "phylum")

merge_phylo <- merge_phyloseq(physeq_genome_filtered_ampl_phylum,physeq_genome_filtered_phylum)
```

```{r}
#create the merge phyloseq object at phylum level
otu1 <- as.data.frame(otu_table(physeq_genome_filtered_phylum))
tax1 <- as.data.frame(tax_table(physeq_genome_filtered_phylum)) %>% select(phylum)
otu1$phylum <- tax1$phylum

otu2 <- as.data.frame(otu_table(physeq_genome_filtered_ampl_phylum))
tax2 <- as.data.frame(tax_table(physeq_genome_filtered_ampl_phylum)) %>% select(phylum)
otu2$phylum <- tax2$phylum

merged_otu <- full_join(otu1, otu2, by = "phylum") %>%
  mutate(across(everything(), ~replace_na(., 0)))%>%
  group_by(phylum) %>%
  summarise(across(everything(), sum, na.rm = TRUE))
merged_otu_mat <- as.matrix(merged_otu %>% select(-phylum))
rownames(merged_otu_mat) <- merged_otu$phylum
otu_table_merged <- otu_table(merged_otu_mat, taxa_are_rows = TRUE)

merged_physeq_phylum <- phyloseq(otu_table_merged, sample_data(merge_phylo))

```

#### Eptesicus
```{r}
physeq_eb <- subset_samples(merged_physeq_phylum, Species == "Eb")
physeq_eb <- prune_taxa(taxa_sums(physeq_eb)>0, physeq_eb)
```

```{r}
ancom_rand_output_struc_eb = ancombc2(data = physeq_eb, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Structural zeros***
```{r}
structural_zeros_eb <- ancom_rand_output_struc_eb$zero_ind %>% 
  as.data.frame() %>% 
  column_to_rownames(., "taxon")

rownames(structural_zeros_eb)[apply(structural_zeros_eb, 1, any)]
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_eb$res %>% 
  filter(p_methodshotgun<0.05)
```

#### Pipistrellus
```{r}
physeq_pk <- subset_samples(merged_physeq_phylum, Species == "Pk")
physeq_pk <- prune_taxa(taxa_sums(physeq_pk)>0, physeq_pk)
```

```{r}
ancom_rand_output_struc_pk = ancombc2(data = physeq_pk, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Structural zeros***
```{r}
structural_zeros_pk <- ancom_rand_output_struc_pk$zero_ind %>% 
  as.data.frame() %>% 
  column_to_rownames(., "taxon")

rownames(structural_zeros_pk)[apply(structural_zeros_pk, 1, any)]
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_pk$res %>% 
  filter(p_methodshotgun<0.05)
```


#### Hypsugo (not working random effect)
```{r}
physeq_ha <- subset_samples(merged_physeq_phylum, Species == "Ha")
physeq_ha <- prune_taxa(taxa_sums(physeq_ha)>0, physeq_ha)
```

```{r}
ancom_rand_output_struc_ha = ancombc2(data = physeq_ha, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
#                  rand_formula = "(1|individual)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
 #                 lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Structural zeros***
```{r}
structural_zeros_ha <- ancom_rand_output_struc_ha$zero_ind %>% 
  as.data.frame() %>% 
  column_to_rownames(., "taxon")

rownames(structural_zeros_ha)[apply(structural_zeros_ha, 1, any)]
```

***Phyla differentially abundant***
```{r}
ancom_rand_output_struc_ha$res %>% 
  filter(p_methodshotgun<0.05)
```
