# Metabarcoding selected filtering based on alpha diversity (t3_p10)

```{r knitr_options_comp, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=7, fig.fullwidth=TRUE,
fig.path='Figures/', warning=FALSE,
message=FALSE)
set.seed(24022025)
```

```{r load_both, warning=FALSE, comments="", message=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
load("resources/amplicon/selected_filtering.Rdata")
species_color <- c("#e5bd5b", "#6b7398", "#76b183")
```

**Number of MAGs**
```{r mag_number_sel, comment="", message=FALSE, warning=FALSE}
genome_metadata_amplicon %>% 
  nrow()%>% 
  cat()
```

**Number of Archaea phyla**
```{r arch_phyla_meta_sel, comment="", message=FALSE, warning=FALSE}
genome_metadata_amplicon %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

**Number of Bacteria phyla**
```{r bact_meta_sel, comment="", message=FALSE, warning=FALSE}
genome_metadata_amplicon %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()%>% 
  cat()
```

## Taxonomy
```{r taxonomy_barplot_selec_filt, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt_amplicon %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata_amplicon, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata_amplicon, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_nested( ~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 10),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 1))
```

**Grouping low-abundance bacteria**
```{r group_plot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
p1 <- genome_counts_filt %>%
  mutate_at(vars(-genome), ~ . / sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>%
group_by(sample, phylum) %>%
  mutate(total_abundance = sum(count)) %>%
  ungroup() %>%
  mutate(phylum = if_else(total_abundance < 0.01, "Other", phylum)) %>%
  group_by(sample, phylum, Species) %>%
  summarise(count = sum(count), .groups = "drop") %>%
   ggplot(aes(
    x = sample,
    y = count,
    fill = phylum,
    group = phylum
  )) + 
  geom_bar(stat = "identity",
           colour = "white",
           linewidth = 0.1) + 
  scale_fill_manual(values = c(phylum_colors, "Other" = "grey50"),
                    drop = FALSE)+
  facet_nested(~ factor(
    Species,
    labels = c("Eb" = "Cnephaeus", "Ha" = "Hypsugo", "Pk" = "Pipistrellus")
  ), scales = "free") + 
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.text = element_text(size = 10),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(
      size = 12,
      lineheight = 0.6,
      face = "bold"
    ),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")+
  guides(fill = guide_legend(ncol = 1))

#ggsave("community_plot_grouped_selected.pdf", plot = p1, width = 12, height = 6)
p1
```
**Phylum relative abundances**
```{r taxonomy_phylum_selected, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt_amplicon %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  left_join(genome_metadata_amplicon, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,Species) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_selected, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    Total_mean = mean(relabun * 100, na.rm = T),
    Total_sd = sd(relabun * 100, na.rm = T),
    Eb_mean = mean(relabun[Species == "Eb"] * 100, na.rm = T),
    Eb_sd = sd(relabun[Species == "Eb"] * 100, na.rm = T),
    Ha_mean = mean(relabun[Species == "Ha"] * 100, na.rm = T),
    Ha_sd = sd(relabun[Species == "Ha"] * 100, na.rm = T),
    Pk_mean = mean(relabun[Species == "Pk"] * 100, na.rm = T),
    Pk_sd = sd(relabun[Species == "Pk"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(Total_mean, 3), "±", round(Total_sd, 3)),
    Cnephaeus = str_c(round(Eb_mean, 3), "±", round(Eb_sd, 3)),
    Hypsugo = str_c(round(Ha_mean, 3), "±", round(Ha_sd, 3)),
    Pipistrellus = str_c(round(Pk_mean, 3), "±", round(Pk_sd, 3))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(phylum,
                Total,
                Cnephaeus,
                Hypsugo,
                Pipistrellus) %>%
  tt()
```
```{r dist_ASv_ampli, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

total_asvs <- data.frame(
  Bat = character(),
  ASVs = numeric(),
  Phylum = numeric(),
  Family = numeric(),
  Genus = numeric()
)

preabs_table <- genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata_amplicon[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata_amplicon, by = join_by("genome" == "genome")) %>% 
  filter(domain=="Bacteria")

phylum <- preabs_table %>%
  distinct(phylum)

family <- preabs_table %>%
  distinct(phylum, class, order, family)

genus <- preabs_table %>%
  distinct(phylum, class, order, family, genus)

total_asvs <- rbind(
  total_asvs,
  data.frame(
    Bat = "Total",
    ASVs = nrow(preabs_table),
    Phylum = nrow(phylum),
    Family = nrow(family),
    Genus = nrow(genus)
  )
)

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  phylum <- preabs_table %>%
    select({{bat}}, phylum) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum)
  
  family <- preabs_table %>%
    select({{bat}}, phylum, class, order, family) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
    distinct(phylum, class, order, family, genus)
  
  total_asvs <- rbind(
    total_asvs,
    data.frame(
      Bat = bat,
      ASVs = nrow(number),
      Phylum = nrow(phylum),
      Family = nrow(family),
      Genus = nrow(genus)
    )
  )
}
```

```{r total_annot_ampli_species, warning=FALSE, comments="", message=FALSE}
bats = c("Eb", "Pk", "Ha")

no_annotation <- data.frame(Bat = character(),
                            No_genus = numeric(),
                            No_species = numeric())

preabs_table <- genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata_amplicon[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("genome")  %>%
  left_join(genome_metadata_amplicon, by = join_by("genome" == "genome")) %>% 
  filter(domain=="Bacteria")

genus <- preabs_table  %>%
  filter(is.na(genus))

species <- preabs_table  %>%
  filter(is.na(species))

no_annotation <- rbind(no_annotation,
                       data.frame(
                         Bat = "Total",
                         No_genus = nrow(genus),
                         No_species = nrow(species)
                       ))

for (bat in bats) {
  number <- preabs_table %>%
    select({{bat}}) %>%
    filter(. >= 1)
  
  genus <- preabs_table %>%
    select({{bat}}, phylum, class, order, family, genus) %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(genus))
  
  species <- preabs_table %>%
    filter(!!sym(bat) >= 1) %>%
  filter(is.na(species))
  
  no_annotation <- rbind(no_annotation,
                         data.frame(
                           Bat = bat,
                           No_genus = nrow(genus),
                           No_species = nrow(species)
                         ))
}
```

**Total percentage of ASVs without genus-level annotation**
```{r genon_percet_ampli, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata_amplicon %>%
  filter(domain=="Bacteria") %>%
  filter(is.na(genus)) %>% 
  summarize(ASV_nogenera = n()) %>%
  pull()
nasvs <- total_asvs %>%
  filter(Bat == "Total") %>%
  select(ASVs) %>%
  pull()
perct <- nongenera * 100 / nasvs
cat(perct)
```

**Percentage of ASVs without genus-level annotation by phylum**
```{r nongene_phylum_ampli, comment="", message=FALSE, warning=FALSE}
total_asv_phylum <- genome_metadata_amplicon %>%
  filter(domain=="Bacteria") %>%
  group_by(phylum) %>%
  summarize(Total_ASVs = n())
genome_metadata_amplicon %>%
  filter(is.na(genus)) %>% 
  group_by(phylum) %>%
  summarize(ASVs_nogenus = n()) %>%
  left_join(total_asv_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Percentage_nogenus = 100 * ASVs_nogenus / Total_ASVs) %>%
  tt()
```

**Number of bacterial species**
```{r phyla_ampli_bacteria, comment="", message=FALSE, warning=FALSE}
genome_metadata_amplicon %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(species) %>%
  unique() %>%
  drop_na() %>% 
  pull() %>%
  length() %>% 
  cat()
```

**Total percentage of ASVs without species-level annotation**
```{r sp_percet_ampli_filt, comment="", message=FALSE, warning=FALSE}
nonspecies <- genome_metadata_amplicon %>%
  filter(domain=="Bacteria") %>%
  filter(is.na(species)) %>%
  summarize(ASV_nospecies = n()) %>%
  pull()
perct <- nonspecies * 100 / nasvs
cat(perct)
```

**ASVs without species-level annotation**
```{r nonspe_phylum_ampli_filt, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata_amplicon %>%
  filter(domain=="Bacteria") %>%
  group_by(phylum) %>%
  summarize(ASVs_total = n())

genome_metadata_amplicon %>%
  filter(domain=="Bacteria") %>%
  filter(is.na(species)) %>%
  group_by(phylum) %>%
  summarize(ASVs_nospecies = n()) %>%
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>%
  mutate(Species_annotated = ASVs_total - ASVs_nospecies) %>%
  mutate(Percentage_nospecies = 100 * ASVs_nospecies / ASVs_total) %>%
  mutate(Percentage_species = 100 - 100 * ASVs_nospecies / ASVs_total) %>%
  tt()
```

### Summary table

```{r single_sp_ampli_filt, comment="", message=FALSE, warning=FALSE}
bats = c("Eb", "Pk", "Ha")

single_sp <- data.frame(Bat = character(), Single_species = numeric())

bacteria <- genome_metadata_amplicon %>% 
  filter(domain=="Bacteria")

table_upset_analysis <- genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  filter(genome %in% bacteria$genome) %>% 
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata_amplicon[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample")

unique_all <- table_upset_analysis %>%
  filter(rowSums(across(Eb:Pk)) == 1)

single_sp <- rbind(single_sp, data.frame(Bat = "Total",
                                         Single_species = nrow(unique_all)))
  
for (bat in bats) {
  unique <- table_upset_analysis %>%
    filter(rowSums(across(Eb:Pk)) == 1) %>%
    column_to_rownames(., "sample") %>% 
    select(bat) %>%
    filter(. > 0) %>%
    nrow()
  single_sp <- rbind(single_sp, data.frame(Bat = bat, Single_species = unique))
}
```

```{r single_ind_ampli_filt, comment="", message=FALSE, warning=FALSE}
single_ind <- data.frame(Bat = character(), Single_individual = numeric())

freq_table <- genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  filter(genome %in% bacteria$genome) %>%
  column_to_rownames("genome") %>%
  mutate(across(everything(), ~ as.integer(. > 0))) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata_amplicon[c("sample", "Species")], by = "sample") %>%
  group_by(Species) %>%
  summarize(across(-sample, sum), .groups = "drop") %>%
  column_to_rownames("Species") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("asv")

singleton_filt <- freq_table %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(-asv))) %>%
  filter(row_sum == 1) %>%
  column_to_rownames(var = "asv")  %>%
  filter(row_sum == 1)

single_ind <- rbind(single_ind, data.frame(
  Bat = "Total",
  Single_individual = nrow(singleton_filt)
))
     
for (bat in bats) {
  singleton_filt <- freq_table %>%
    rowwise() %>%
    mutate(row_sum = sum(c_across(-asv))) %>%
    filter(row_sum == 1) %>%
    column_to_rownames(var = "asv")  %>%
    select(bat) %>%
    filter(. == 1)
  
  single_ind <- rbind(single_ind, data.frame(
    Bat = bat,
    Single_individual = nrow(singleton_filt) 
  ))
}
```

```{r merge_ampli_filt, comment="", message=FALSE, warning=FALSE}
summary_table_ampli <- total_asvs %>%
  left_join(., no_annotation, by = "Bat") %>%
  left_join(., single_ind, by = "Bat") %>%
  left_join(., single_sp, by = "Bat") %>%
  mutate(Bat = recode(Bat, 
                      Eb = "Cnephaeus bottae",
                      Ha = "Hypsugo ariel",
                      Pk = "Pipistrellus kuhlii"))
summary_table_ampli %>% 
  tt()
```

```{r summary_table_ampli_sel, comment="", message=FALSE, warning=FALSE}
total_asv <- nrow(genome_metadata_amplicon)
summary_table_ampli %>% 
  select(-Phylum,-Family, -Genus) %>% 
  rowwise() %>% 
  mutate(ASV_perc=round(ASVs*100/total_asv, 2))%>% 
  mutate(No_genus_perc=round(No_genus*100/ASVs, 2))%>% 
  mutate(No_species_perc=round(No_species*100/ASVs, 2)) %>% 
  mutate(Single_individual_perc=round(Single_individual*100/ASVs, 2))%>% 
  mutate(Single_species_perc=round(Single_species*100/ASVs, 2)) %>% 
  mutate(Single_individual_per_unique=round(Single_individual*100/Single_species, 2)) %>% 
  select(1,7:12) %>% tt()
```

**Shared phyla**
```{r}
meta_phylum <- genome_metadata %>% 
  distinct(domain,phylum)

genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(phylum %in% meta_phylum$phylum)
```

**Only in metagenomic data**
```{r}
ampli_phylum <- genome_metadata_amplicon %>% 
  distinct(domain,phylum)

genome_metadata %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% genome_metadata_amplicon$phylum)
```

**Only in metabarcoding selected data**
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum) %>% 
  filter(!phylum %in% meta_phylum$phylum)
```

### Differences at family level
**Number of families in metagenomic data**
```{r}
load("resources/metagenomics/data_filtered.Rdata")
```

```{r}
cat(genome_metadata %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

**Number of families in metabarcoding selected data**
```{r}
cat(genome_metadata_amplicon %>%
      distinct(domain, phylum, family) %>%
      count() %>%
      pull())
```

**Shared families**
```{r}
meta_family <- genome_metadata %>% 
  distinct(domain,phylum, family)

genome_metadata_amplicon %>% 
  distinct(domain,phylum, family) %>% 
  filter(family %in% meta_family$family)
```

**Only in metagenomic data**
```{r}
ampli_family <- genome_metadata_amplicon %>% 
  distinct(domain,phylum,family)

genome_metadata %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% ampli_family$family)
```

**Only in metabarcoding selected data**
```{r}
genome_metadata_amplicon %>% 
  distinct(domain,phylum,family) %>% 
  filter(!family %in% meta_family$family)
```

### Differences in top ASVs and MAGs
Top 20 MAGs
```{r top_mags, comment="", message=FALSE, warning=FALSE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  mutate(total = rowSums(across(-genome))) %>%
  arrange(desc(total)) %>%
  slice_head(n = 20) %>% 
  select(genome, total) %>% 
  left_join(genome_metadata[1:8])
```

Top 20 ASVs
```{r top_asv, comment="", message=FALSE, warning=FALSE}
genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  mutate(total = rowSums(across(-genome))) %>%
  arrange(desc(total)) %>%
  slice_head(n = 20) %>% 
  select(genome, total) %>% 
  left_join(genome_metadata_amplicon)
```

Phylum level
```{r top_mags_phylum, comment="", message=FALSE, warning=FALSE}
topMAG <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1, 3, 6, 7)], by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum) %>% 
  arrange(desc(total_mag))%>%
  slice_head(n = 10)
```

```{r top_asv_phylum, comment="", message=FALSE, warning=FALSE}
topASV <- genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>% 
  select(-genome) %>% 
  group_by(phylum) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 10)

table <- full_join(topMAG,topASV, by="phylum") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
  select(-total_asv, -total_mag) %>% 
  pivot_longer(cols = c(MAG, ASV),
               names_to = "source",
               values_to = "abundance")
```

```{r figure_bubble_phylum_filt, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
ggplot(table, aes(x = source, y = phylum, size = abundance, color = phylum)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(
    name = "Abundance",
    range = c(0.01, 30),   # 0 ensures no bubble at abundance = 0
    breaks = seq(0, 35, by = 5),   # Controls legend labels (your 0–5–10–...–30)
    limits = c(0.01, NA)   # keep full range of values
  )  +
    scale_color_manual(values = phylum_colors) +
    coord_flip()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45,hjust = 1, size=12),
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14, face = "bold"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12))+
  labs(x = "Source", y = "Phylum",
    title = "Metabarcoding with selected threshold")+
  guides(color = "none", size = guide_legend(
      override.aes = list(size = c(2, 4, 6, 8, 10, 12)) # smaller bubbles in legend
    )
  )
```
```{r figure_scat_phylum_filt, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
full_join(topMAG,topASV, by="phylum") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
  ggplot(., aes(x = MAG, y = ASV, color = phylum, label = phylum)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 20) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
  labs(
    x = "MAG abundance",
    y = "ASV abundance",
    color = "Phylum",
    title = "Comparison of MAG and ASV abundances by phylum"
  ) +
  theme_minimal()
```

Family level
```{r family_top_mag, comment="", message=FALSE, warning=FALSE}
topFMAG <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata[c(1:7)], by="genome") %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )%>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_mag = rowSums(across(where(is.numeric)))) %>% 
  select(total_mag, phylum, family) %>% 
  arrange(desc(total_mag)) %>%
  slice_head(n = 20)
```
```{r family_top_asv, echo=FALSE}
topFASV <-genome_counts_filt_amplicon %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  left_join(genome_metadata_amplicon, by="genome") %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  ) %>% 
  select(-genome) %>% 
  group_by(phylum, family) %>% 
  summarize(across(where(is.numeric), sum), .groups = "drop") %>% 
  mutate(total_asv = rowSums(across(where(is.numeric)))) %>% 
  select(total_asv, phylum, family) %>% 
  arrange(desc(total_asv)) %>%
  slice_head(n = 20)
```

```{r figure_scat_family_top, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=12, fig.fullwidth=TRUE}
full_join(topFMAG,topFASV, by="family") %>% 
    mutate(ASV = replace_na(total_asv, 0)) %>% 
    mutate(MAG = replace_na(total_mag, 0)) %>% 
    mutate(phylum_mag = coalesce(phylum.x, phylum.y))%>%
  ggplot(., aes(x = MAG, y = ASV, color = phylum_mag, label = family)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text_repel(max.overlaps = 10) +
  geom_smooth(method = "lm", se = FALSE, color = "grey50") +
      scale_color_manual(values = phylum_colors) +
  labs(
    x = "MAG abundance",
    y = "ASV abundance",
    color = "Phylum",
    title = "Metabarcoding with selected filtering"
  ) +
  theme_minimal()
```

## Alpha Diversity

### ASV level
```{r alpha_cal_filt_selected, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt_amplicon %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

metabarcoding_selected_alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>% 
  mutate(method="Metabarcoding_selected")
```

**Correlation between diversity and estimate microbial fraction**
```{r fraction_selected, comment="", message=FALSE, warning=FALSE}
fraction_ampli <- microbial_fraction %>% 
  select(sample, estimated_microbial_fraction) %>% 
  right_join(metabarcoding_selected_alpha_div, by="sample") %>% 
  left_join(sample_metadata_amplicon, by="sample") 

cor.test(fraction_ampli$estimated_microbial_fraction, fraction_ampli$richness, method = "spearman")
cor.test(fraction_ampli$estimated_microbial_fraction, fraction_ampli$neutral, method = "spearman")
```

```{r fraction_ampli_fig_selected, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
p1_ampli <- ggplot(fraction_ampli, aes(x = estimated_microbial_fraction, y = richness)) +
  geom_point(size = 3, color = "#2E86AB") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Richness", title= "Correlation between estimated microbial fraction and diversity")

p2_ampli <- ggplot(fraction_ampli, aes(x = estimated_microbial_fraction, y = neutral)) +
  geom_point(size = 3, color = "#F39C12") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_classic() +
  labs(x = "Read Fraction (%)", y = "Neutral Alpha Diversity")

p1_ampli + p2_ampli 
```
```{r alpha_div_selected, comment="", message=FALSE, warning=FALSE}
richness %>%
  full_join(neutral, by = join_by(sample == sample))  %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

### Different taxonomic level
```{r}
sample_metadata_amplicon <- sample_metadata_amplicon[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli_selec")) %>%
  mutate(method="Metabarcoding_selected")

genome_metadata_amplicon <- genome_metadata_amplicon %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt_amplicon %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli_selec")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_amplicon, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

#### Family level
```{r alpha_family_indiv_fil, comment="", message=FALSE, warning=FALSE}
richness_f_ampli <- family_level_agg_ampli %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_f_ampli <- family_level_agg_ampli %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_f_ampli <- richness_f_ampli %>%
  full_join(neutral_f_ampli, by = join_by(sample == sample))

alpha_div_f_ampli_data <- alpha_div_f_ampli %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_fam_summary_all_ampli_sel, comment="", message=FALSE, warning=FALSE}
alpha_div_f_ampli %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**

```{r rich_MM_general_indiv_fil, comment="", message=FALSE, warning=FALSE}
Model_richness <- MASS::glm.nb(richness ~ Species, data = alpha_div_f_ampli_data,trace=TRUE)
summary(Model_richness)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

**Neutral**
```{r neutral_MM_general_indiv_fil, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_f_ampli_data)
anova(Model_neutral)
emmeans(Model_neutral, pairwise ~ Species)
```

#### Phylum level
```{r alpha_family_indiv, comment="", message=FALSE, warning=FALSE}
richness_p_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_p_ampli <- phylum_level_agg_ampli %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_phylum_ampli <- richness_p_ampli %>%
  full_join(neutral_p_ampli, by = join_by(sample == sample))

alpha_div_phylum_ampli_data <- alpha_div_phylum_ampli %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample))
```

**Diversity values**
```{r alpha_div_phy_summary_all_ampli_sel, comment="", message=FALSE, warning=FALSE}
alpha_div_phylum_ampli %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Eb_mean = mean(value[Species == "Eb"], na.rm = T),
    Eb_sd = sd(value[Species == "Eb"], na.rm = T),
    Ha_mean = mean(value[Species == "Ha"], na.rm = T),
    Ha_sd = sd(value[Species == "Ha"], na.rm = T),
    Pk_mean = mean(value[Species == "Pk"], na.rm = T),
    Pk_sd = sd(value[Species == "Pk"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    Cnephaeus = str_c(round(Eb_mean, 2), "±", round(Eb_sd, 2)),
    Hypsugo = str_c(round(Ha_mean, 2), "±", round(Ha_sd, 2)),
    Pipistrellus = str_c(round(Pk_mean, 2), "±", round(Pk_sd, 2))
  ) %>%
  arrange(-Eb_mean) %>%
  dplyr::select(alpha, Total, Cnephaeus, Pipistrellus, Hypsugo) %>%
  tt()
```

**Richness**
```{r rich_MM_general_indiv, comment="", message=FALSE, warning=FALSE}
Model_richness <- MASS::glm.nb(richness ~ Species, data = alpha_div_phylum_ampli_data,trace=TRUE)
summary(Model_richness)
anova(Model_richness)
emmeans(Model_richness, pairwise ~ Species)
```

**Neutral**
```{r neutral_MM_general_indiv, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lm(formula = neutral ~ Species, data = alpha_div_phylum_ampli_data)
anova(Model_neutral)
```

### Comparison (metabarcoding selected vs metagenomics)

```{r}
load("resources/alpha_together.Rdata")
```

Merge the table
```{r}
merged_table_family_methods <- full_join(family_level_agg_ampli,merged_table, 
          by = c("family"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

merged_table_phylum_methods <- full_join(phylum_level_agg_ampli,merged_table_phylum, 
          by = c("phylum"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

all_samples <- rbind(all_samples, sample_metadata_amplicon)
```

#### Family level
```{r alpha_family, comment="", message=FALSE, warning=FALSE}
richness <- merged_table_family_methods %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- merged_table_family_methods %>%
  column_to_rownames(var = "family") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_family <- richness %>%
  full_join(neutral, by = join_by(sample == sample))

alpha_div_family_data <- alpha_div_family %>%
  left_join(all_samples, by = join_by(sample == sample))

alpha_div_family_data_filt <- alpha_div_family_data %>% 
  filter(method!="Metabarcoding_standard") %>% 
  mutate(sampleid = str_remove(sample, "_ampli_selec"))
```

**Richness**
```{r rich_MM_ampli_sel_species_family, comment="", message=FALSE, warning=FALSE}
Model_richness_random <- glmer.nb(richness ~ Species*method+(1|sampleid), data = alpha_div_family_data_filt)
summary(Model_richness_random)
anova(Model_richness_random)
emmeans(Model_richness_random, pairwise ~ Species)
emmeans(Model_richness_random, pairwise ~ method | Species)
```

```{r exapl_fam_rich, comment="", message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
exp(2.639) ≈ 14
Baseline richness ≈ 14 taxa

Main effects:
Species
SpeciesHa = -0.289 (ns)
SpeciesPk = -0.415 (borderline)

  Compared to baseline species (amplicon):
    Ha → slightly lower richness (not significant)
    Pk → somewhat lower richness (~ borderline)
 Species emmean     SE  df asymp.LCL asymp.UCL
 Eb        2.47 0.1230 Inf      2.23      2.71
 Ha        1.70 0.1060 Inf      1.49      1.91
 Pk        1.99 0.0961 Inf      1.81      2.18
   contrast estimate    SE  df z.ratio p.value
 Eb - Ha     0.770 0.161 Inf   4.775  <.0001
 Eb - Pk     0.475 0.154 Inf   3.079  0.0059
 Ha - Pk    -0.295 0.142 Inf  -2.080  0.0939
 
 Eb>Pk≈Ha
    
methodMetagenomics = -0.340 (ns)
  For the baseline species only:
    Metagenomics gives lower richness than amplicon
    But not statistically significant

The IMPORTANT part: Interaction
  SpeciesHa:methodMetagenomics = -0.960 (p = 0.0029)
  This is the key biological result.
  The effect of metagenomics vs amplicon is different in Species Ha, and specifically much more negative.

Baseline species

Effect of metagenomics:
  −0.340
  Slight (non-significant) decrease in richness

Species Ha
  −0.340+(−0.960)=−1.300
  Strong decrease in richness with metagenomics
  exp(-1.30) ≈ 0.27
  ~73% lower richness with metagenomics vs amplicon
  Statistically significant

Species Pk
  −0.340+(−0.119)=−0.459
  Mild decrease
  exp(-0.459) ≈ 0.63
  ~37% lower richness
  Not significant    
```

**Neutral**
```{r neutral_MM_ampli_sel_species_family, comment="", message=FALSE, warning=FALSE}
Model_neutral <- lme(fixed = neutral ~ Species*method, data = alpha_div_family_data_filt,
               random = ~ 1 | sampleid)
anova(Model_neutral)
summary(Model_neutral)
emmeans(Model_neutral, pairwise ~ method | Species)
```
```{r exapl_fam_neutral, comment="", message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
Species         p = 0.080   (marginal) Species alone is only weakly significant
method          p = 0.0197  (*) Method matters overall
Species:method  p = 0.045   (*) Interaction is significant
The effect of sequencing method on neutral depends on the species.
Why ANOVA is significant but coefficients are not

This is important:

👉 The ANOVA says:

“There is some interaction effect overall”

But:

Individual coefficients are noisy (large SE)
No single species contrast is strongly significant

👉 This happens when:

Effects are distributed across groups
Variability is high
```

#### Phylum level
```{r alpha_phylum_sel, comment="", message=FALSE, warning=FALSE}
richness_phylum <- merged_table_phylum_methods %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral_phylum <- merged_table_phylum_methods %>%
  column_to_rownames(var = "phylum") %>%
  dplyr::select(where( ~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

alpha_div_phylum <- richness_phylum %>%
  full_join(neutral_phylum, by = join_by(sample == sample))

alpha_div_phylum_data <- alpha_div_phylum %>%
  left_join(all_samples, by = join_by(sample == sample))

alpha_div_phylum_data_filt <- alpha_div_phylum_data %>% 
  filter(method!="Metabarcoding_standard") %>% 
  mutate(sampleid = str_remove(sample, "_ampli_selec"))
```

**Richness**
```{r rich_alpha_phylum, comment="", message=FALSE, warning=FALSE}
Model_richness_phylum_random <- glmer.nb(richness ~ Species*method+(1|sampleid), data = alpha_div_phylum_data_filt)
summary(Model_richness_phylum_random)
anova(Model_richness_phylum_random)
emmeans(Model_richness_phylum_random, pairwise ~ Species)
emmeans(Model_richness_phylum_random, pairwise ~ method | Species)
```

```{r expla1, comment="", message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
The random effect (sample) is essentially doing nothing.

Variance = ~0
Model behaves like a regular GLM

---
Intercept = 1.74
exp(1.74) ≈ 5.7

Baseline richness ≈ 5–6 phyla

--
  
Species effects (strong!)
SpeciesHa = -0.73  (***)  
SpeciesPk = -0.48  (**)

Compared to baseline species:

Ha → significantly lower richness
Pk → significantly lower richness

Strong biological signal:

Species differ in phylum-level richness

--
Emmeans:  
First, the warning (important but not a problem)

“Results may be misleading due to involvement in interactions”

This appears because the model includes Species × method.

Meaning:

These are averaged across methods
But since you know there is some interaction, results should be interpreted as overall trends, not species-specific method effects

Still valid for the question (Ha vs Pk)

Estimated means:
Eb ≈ 5.39  
Ha ≈ 1.82  
Pk ≈ 3.30

Clear ranking:

Eb > Pk > Ha

---
Method effect (baseline species only)
methodMetagenomics = -0.11 (ns)
For baseline species:

No meaningful difference between methods
---
 Interaction significant
SpeciesHa:methodMetagenomics = -0.70 (p = 0.019)
Baseline species

Effect of metagenomics:
−0.11
No real change

Species Ha
−0.11+(−0.70)=−0.81
exp(-0.81) ≈ 0.44
 ~56% lower richness with metagenomics
Significant

Species Pk:
−0.11+(−0.02)=−0.13
exp(-0.13) ≈ 0.88
Small decrease
Not significant

```

**Neutral**
```{r neutral_alpha_phylum, comment="", message=FALSE, warning=FALSE}
Model_neutral_phylum <- lme(fixed = neutral ~ Species*method, data = alpha_div_phylum_data_filt,
               random = ~ 1 | sampleid)
anova(Model_neutral_phylum)
summary(Model_neutral_phylum)
emmeans(Model_neutral_phylum, pairwise ~ Species)
emmeans(Model_neutral_phylum, pairwise ~ method | Species)
```
```{r expla2, comment="", message=FALSE, warning=FALSE, eval=FALSE, echo=FALSE}
Estimated means:
Eb ≈ 2.42
Pk ≈ 2.10
Ha ≈ 1.39

Clear ranking:

Eb ≈ Pk > Ha

Ha is consistently lowest
Eb and Pk are not different
```

### Alpha diversity figure (three methods)

```{r fig_family, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=16, fig.fullwidth=TRUE}
alpha_div_family_data %>%
  select(-sample,-Life_stage, -Sex) %>%
  pivot_longer(
    cols = -c(method, Species),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("richness", "neutral"),
      labels = c("Richness", "Neutral")
    ),
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii")
      ),
    method = factor(
      method,
      levels = c("Metabarcoding_standard", "Metabarcoding_selected", "Metagenomics"),
      labels = c("Metabarcoding Standard", "Metabarcoding Selected", "Metagenomics")
  )) %>%   
  ggplot(aes(y = value, x = method, color = Species, fill = Species)) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA,
 #   show.legend = FALSE,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
    size = 1,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  scale_fill_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = 0.1, color = "grey"),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 14),
    strip.text = element_text(size = 14, lineheight = 0.6, face = "bold")
  ) +
  labs(
#    x = "Methodological approach",
    y = "Alpha diversity at family level",
    fill = "Species"
  )
```

```{r gif_phylum_sel, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=16, fig.fullwidth=TRUE}
alpha_div_phylum_data %>%
  select(-sample,-Life_stage, -Sex) %>%
  pivot_longer(
    cols = -c(method, Species),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("richness", "neutral"),
      labels = c("Richness", "Neutral")
    ),
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii")
      ),
    method = factor(
      method,
      levels = c("Metabarcoding_standard", "Metabarcoding_selected", "Metagenomics"),
      labels = c("Metabarcoding Standard", "Metabarcoding Selected", "Metagenomics")
  )) %>%   
  ggplot(aes(y = value, x = method, color = Species, fill = Species)) +
  geom_boxplot(
    width = 0.4,
    alpha = 0.5,
    outlier.shape = NA,
 #   show.legend = FALSE,
    position = position_dodge(width = 0.75)
  ) +
  geom_jitter(
    position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75),
    size = 1,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  facet_wrap(~metric, scales = "free_y") +
  scale_color_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii"))))+
  scale_fill_manual(values=species_color,  labels = c(expression(italic("C. bottae")),
             expression(italic("H. ariel")),
             expression(italic("P. kuhlii")))) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    panel.grid.minor.x = element_line(size = 0.1, color = "grey"),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 14),
    strip.text = element_text(size = 14, lineheight = 0.6, face = "bold")
  ) +
  labs(
#    x = "Methodological approach",
    y = "Alpha diversity at phylum level",
    fill = "Species"
  )
```

## Beta diversity

### At different taxonomic level
```{r beta_div_selected_family_merged, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- family_level_agg_ampli %>%
  column_to_rownames(., "family") %>% 
  hillpair(., q = 0)

beta_q1n <- family_level_agg_ampli %>%
  column_to_rownames(., "family") %>% 
  hillpair(., q = 1)

beta_q0n_phylum <- phylum_level_agg_ampli %>%
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 0)

beta_q1n_phylum <- phylum_level_agg_ampli %>%
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 1)
```

```{r save_beta_sele, comment="", message=FALSE, warning=FALSE, eval=FALSE}
save(beta_q0n,
     beta_q1n,
     beta_q0n_phylum,
     beta_q1n_phylum,
     file = "resources/beta_tax_sele.Rdata")
```

```{r load_beta_selec, comment="", message=FALSE, warning=FALSE}
load("resources/beta_tax_sele.Rdata")
```

#### Family
**Richness**
```{r}
adonis2(
  beta_q0n$S ~ Species,
  data = sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n$S,
  sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q0n$S)) %>%
    arrange(match(sample, labels(beta_q0n$S))) %>%
    pull(Species),
  perm = 999) %>%
  tt()
```

**Neutral**
```{r}
adonis2(
  beta_q1n$S ~ Species,
  data = sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q1n$S)) %>%
    arrange(match(sample, labels(beta_q1n$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q1n$S,
  sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q1n$S)) %>%
    arrange(match(sample, labels(beta_q1n$S))) %>%
    pull(Species),
  perm = 999) %>%
  tt()
```

#### Phylum
**Richness**
```{r}
adonis2(
  beta_q0n_phylum$S ~ Species,
  by = "terms",
  data = sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q0n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q0n_phylum$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()

pairwise.adonis(
  beta_q0n_phylum$S,
  sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q0n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q0n_phylum$S))) %>%
    pull(Species),
  perm = 999) %>%
  tt()
```

**Neutral**
```{r}
adonis2(
  beta_q1n_phylum$S ~ Species,
  data = sample_metadata_amplicon %>%
    filter(sample %in% labels(beta_q1n_phylum$S)) %>%
    arrange(match(sample, labels(beta_q1n_phylum$S))),
  permutations = 999
) %>%
  broom::tidy() %>%
  tt()
```

### Comparison between metabarcoding selected and metagenomics

```{r beta_div_together_family_merged, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- merged_table_family_methods %>%
  select(-H19, -H20) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 0)

beta_q1n <- merged_table_family_methods %>%
  select(-H19) %>% 
  column_to_rownames(., "family") %>% 
  hillpair(., q = 1)

beta_q0n_phylum <- merged_table_phylum_methods %>%
  select(-H19, -H20) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 0)

beta_q1n_phylum <- merged_table_phylum_methods %>%
  select(-H19) %>% 
  column_to_rownames(., "phylum") %>% 
  hillpair(., q = 1)
```

```{r save_beta, comment="", message=FALSE, warning=FALSE, eval=FALSE}
save(beta_q0n,
     beta_q1n,
     beta_q0n_phylum,
     beta_q1n_phylum,
     file = "resources/beta_tax.Rdata")
```

```{r load_beta, comment="", message=FALSE, warning=FALSE}
load("resources/beta_tax.Rdata")
methods_colors <- c("#d15aa9","#d4b9ed", "#c7c5c6")
```

#### Family level
**Richness**
```{r permanova_rich_both_fig_family_fil, comment="", message=FALSE, warning=FALSE}
m <- as.matrix(beta_q0n$S)
samples <- rownames(m)
to_remove <- grepl("_ampli$", samples) & !grepl("_ampli_Select$", samples)
m_filtered <- m[!to_remove, !to_remove]
beta_q0n_S_filtered <- as.dist(m_filtered)

samples_sel <- all_samples %>%
  mutate(sampleid = str_remove(sample, "_ampli_selec"))

samples_q0 <- samples_sel %>% filter(sample %in% labels(beta_q0n_S_filtered)) %>% arrange(match(sample, labels(beta_q0n_S_filtered)))

betadisper(beta_q0n_S_filtered, samples_q0 %>% pull(Species)) %>% permutest(.)

adonis2(beta_q0n_S_filtered ~ Species*method, data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q0n_S_filtered ~ Species+method, by = "margin", data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q0n_S_filtered ~ Species*method, by = "margin", data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()
```

**Neutral**
```{r permanova_neutral_tog_Family_fil, comment="", message=FALSE, warning=FALSE}
m <- as.matrix(beta_q1n$S)
samples <- rownames(m)
to_remove <- grepl("_ampli$", samples) & !grepl("_ampli_Select$", samples)
m_filtered <- m[!to_remove, !to_remove]
beta_q1n_S_filtered <- as.dist(m_filtered)

samples_q1 <- samples_sel %>% filter(sample %in% labels(beta_q1n_S_filtered)) %>% arrange(match(sample, labels(beta_q1n_S_filtered)))

betadisper(beta_q1n_S_filtered, samples_q1 %>% pull(Species)) %>% permutest(.)

adonis2(beta_q1n_S_filtered ~ Species*method, data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q1n_S_filtered ~ Species+method, by = "margin", data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q1n_S_filtered ~ Species*method, by = "margin", data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()
```

#### Phylum level
**Richness**
```{r permanova_rich_both_phylum_fig2_fil, comment="", message=FALSE, warning=FALSE}
m <- as.matrix(beta_q0n_phylum$S)
samples <- rownames(m)
to_remove <- grepl("_ampli$", samples) & !grepl("_ampli_Select$", samples)
m_filtered <- m[!to_remove, !to_remove]
beta_q0n_phylum_S_filtered <- as.dist(m_filtered)

samples_q0 <- samples_sel %>% filter(sample %in% labels(beta_q0n_phylum_S_filtered)) %>% arrange(match(sample, labels(beta_q0n_phylum_S_filtered)))

betadisper(beta_q0n_phylum_S_filtered, samples_q0 %>% pull(Species)) %>% permutest(.)

adonis2(beta_q0n_phylum_S_filtered ~ Species*method, data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q0n_phylum_S_filtered ~ Species+method, by = "margin", data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q0n_phylum_S_filtered ~ Species*method, by = "margin", data = samples_q0, permutations = 999, strata = samples_q0$sampleid) %>%
  broom::tidy() %>%
  tt()
```

**Neutral**
```{r permanova_neutral_tog_phylum_fil, comment="", message=FALSE, warning=FALSE}
m <- as.matrix(beta_q1n_phylum$S)
samples <- rownames(m)
to_remove <- grepl("_ampli$", samples) & !grepl("_ampli_Select$", samples)
m_filtered <- m[!to_remove, !to_remove]
beta_q1n_phylum_S_filtered <- as.dist(m_filtered)

samples_q1 <- samples_sel %>% filter(sample %in% labels(beta_q1n_phylum_S_filtered)) %>% arrange(match(sample, labels(beta_q1n_phylum_S_filtered)))

betadisper(beta_q1n_phylum_S_filtered, samples_q1 %>% pull(Species)) %>% permutest(.)

adonis2(beta_q1n_phylum_S_filtered ~ Species*method, data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q1n_phylum_S_filtered ~ Species+method, by = "margin", data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()

adonis2(beta_q1n_phylum_S_filtered ~ Species*method, by = "margin", data = samples_q1, permutations = 999, strata = samples_q1$sampleid) %>%
  broom::tidy() %>%
  tt()
```

### Three methodological approaches together
#### Family
**Richness**
```{r beta_richness_Family_three1, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii")),
      method = factor(
      method,
      levels = c("Metabarcoding_standard", "Metabarcoding_selected", "Metagenomics"),
      labels = c("Metabarcoding Standard", "Metabarcoding Selected", "Metagenomics")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(
    shape = "Methodological approach"
  )#+labs(color = "method") +geom_text_repel(aes(label = sample), size=3)
  
```
```{r beta_richness_Family_three2, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(method = factor(method,
      levels = c("Metabarcoding_standard", "Metabarcoding_selected", "Metagenomics"),
      labels = c("Metabarcoding Standard", "Metabarcoding Selected", "Metagenomics")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = method)) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = methods_colors)+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  )
```

**Neutral**
```{r beta_neutral_Family_three1, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii"))
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 3) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(shape = "Methodological approach")
```
```{r beta_neutral_Family_three2, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  filter(sample!="P36") %>% 
  group_by(method) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
    mutate(method = factor(method,
      levels = c("Metabarcoding_standard", "Metabarcoding_selected", "Metagenomics"),
      labels = c("Metabarcoding Standard", "Metabarcoding Selected", "Metagenomics")
      )
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = method)) + 
  geom_point(size = 3) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = methods_colors)+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Methodological approach",shape="Species")
```

#### Phylum level
**Richness**
```{r beta_div1_both_fig_phylum_fil, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q0n_phylum$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method, Species) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    Species = factor(
      Species,
      levels = c("Eb", "Ha", "Pk"),
      labels = c("C. bottae", "H. ariel", "P. kuhlii"))
  ) %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Species, shape=method)) +
  geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = species_color,
                     labels = c(
                       expression(italic("C. bottae")),
                       expression(italic("H. ariel")),
                       expression(italic("P. kuhlii"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(
    shape = "Methodological approach"
  )
  
```

**Neutral**
```{r beta_div1_both_fig3_phylum_fil, comment="", message=FALSE, warning=FALSE, fig.height=8, fig.width=16, fig.fullwidth=TRUE}
beta_q1n_phylum$S %>%
  vegan::metaMDS(.,
                 trymax = 500,
                 k = 2,
                 trace = 0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(all_samples, by = join_by(sample == sample)) %>%
  group_by(method) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = method, shape= Species)) +
  geom_point(size = 3) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2 ), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(values = methods_colors,
                     labels = c(
                       expression(italic("Metabarcoding Standard")),
                       expression(italic("Metabarcoding Selected")),
                       expression(italic("Metagenomics"))
                     ))+
  theme(
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = "Methodological approach",shape="Species")
```

## Differential abundance

### Metabarcoding selected
```{r div_load_data_metab_sel, comment="", message=FALSE, warning=FALSE}
load("resources/amplicon/selected_filtering.Rdata")
```
```{r zero_phylo1_metab_sel, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata_amplicon %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt_amplicon %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_taxonomy <- genome_metadata_amplicon %>% 
  filter(genome %in% rownames(phylo_genome)) %>% 
  column_to_rownames("genome") %>% 
  dplyr::select(domain,phylum,class,order,family,genus,species) %>% 
  as.matrix() %>% 
  tax_table()

physeq_genome_metab_sel <- phyloseq(phylo_genome, phylo_samples, phylo_taxonomy)
```

```{r ancom_rand_phylum_metab_sel, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_phylum_metab_sel = ancombc2(
  data = physeq_genome_metab_sel,
  assay_name = "counts",
  tax_level = "phylum",
  fix_formula = "Species",
  p_adj_method = "holm",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "Species",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 20,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)
```

#### Structural zeros

***Total number of structural zeros***
```{r}
# Extract structural zeros
structural_zeros <- ancom_rand_output_phylum_metab_sel$zero_ind

# Convert to a data frame for better readability
structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

# Identify in which groups each taxon is a structural zero
structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("Eb"=1) %>% 
  rename("Ha"=2) %>% 
  rename("Pk"=3) 

# Total number of structural zeros
nrow(structural_zero_groups)
```

#### Enriched data
```{r  labeling_phyl_metab_sel, comment="", message=FALSE, warning=FALSE}
res_pair_phylum <- ancom_rand_output_phylum_metab_sel$res_pair
# Get log-fold change and q-value columns
lfc_cols <- grep("^lfc_", colnames(res_pair_phylum), value = TRUE)
q_cols   <- gsub("^lfc_", "q_", lfc_cols)

# Build long table by pairing lfc and q for each contrast
plot_data_phylum <- lfc_cols %>%
  setNames(nm = .) %>%
  purrr::imap_dfr(function(lfc_col, name) {
    q_col <- gsub("^lfc_", "q_", lfc_col)
    
    res_pair_phylum %>%
      select(taxon, !!lfc_col, !!q_col) %>%
      rename(
        lfc = !!lfc_col,
        q   = !!q_col
      ) %>%
      mutate(contrast_key = name)
  }) %>%
  filter(!is.na(lfc), !is.na(q), q < 0.05) %>%
  mutate(
    # Manually re-label contrasts for clarity
    contrast = case_when(
      contrast_key == "lfc_SpeciesHa" ~ "H. ariel vs C. bottae",
      contrast_key == "lfc_SpeciesPk" ~ "P. kuhlii vs C. bottae",
      contrast_key == "lfc_SpeciesPk_SpeciesHa" ~ "P. kuhlii vs H. ariel",
      TRUE ~ contrast_key
    ),
    direction = ifelse(lfc > 0, paste("Enriched in", sub(" vs.*", "", contrast)),
                             paste("Enriched in", sub(".*vs ", "", contrast)))
  )
plot_data_phylum %>% select(-contrast_key)
```

### Differences with metagenomics at phylum level across species

```{r}
load("resources/amplicon/selected_filtering.Rdata")

sample_metadata_amplicon <- sample_metadata_amplicon[c(1,9,10,13)] %>%
  mutate(sample = paste0(sample, "_ampli")) %>%
  mutate(method="Metabarcoding")

genome_metadata_ampli <- genome_metadata_amplicon %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(is.na(genus),
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax_ampli <- genome_counts_filt_amplicon %>% 
  column_to_rownames(., "genome")%>%
  rename_with(~ paste0(., "_ampli")) %>%
  rownames_to_column(., "genome")%>%
  left_join(genome_metadata_ampli, by = "genome")

family_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame() %>% 
  select(-c(phylum,class))

phylum_level_agg_ampli <- genome_tax_ampli %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
    as.data.frame()
```

```{r load_data_beta_comparison_sel}
load("resources/metagenomics/data_filtered.Rdata")

genome_metadata <- genome_metadata %>%
  mutate(
    class  = if_else(str_trim(class) == "", paste0(phylum, "_Unclassified"), class),
    order  = if_else(str_trim(order) == "",
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(str_trim(family) == "",
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(str_trim(genus) == "",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus)
  )

genome_tax <- genome_counts_filt %>% 
  left_join(genome_metadata[c(1,3,4,6)], by = "genome")

# Aggregate counts at the Family level
family_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum, class, family) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class))

# Aggregate counts at the Phylum level
phylum_level_agg <- genome_tax %>%
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() 
```

```{r}
merged_table <- full_join(family_level_agg_ampli,family_level_agg, 
          by = c("family"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

merged_table_phylum <- full_join(phylum_level_agg_ampli,phylum_level_agg, 
          by = c("phylum"))%>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))


sample_shot <- sample_metadata %>% 
  select(-c(Habitat,Longitude, Latitude, Region))%>%
  mutate(method="Metagenomics")

all_samples <- rbind(sample_shot, sample_metadata_amplicon)

```
```{r}
phylo_genome <- merged_table_phylum %>%
  column_to_rownames("phylum") %>% 
  otu_table(., taxa_are_rows = TRUE)

phylo_samples <- all_samples%>% 
  mutate(individual=sample)%>%
  column_to_rownames("sample") %>% 
  sample_data()

merged_physeq_phylum <- phyloseq(phylo_genome, phylo_samples)
```
```{r}
ancom_rand_output_struc_approach_rand = ancombc2(data = merged_physeq_phylum, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
    #              rand_formula = "(1|Species)",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
        #          lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Enriched phyla***
```{r}
ancom_rand_output_struc_approach_rand$res%>%
  filter(q_methodMetagenomics < 0.05) %>%
  tt()
```

### Differences with metagenomics at phylum level within each species

#### Cnephaeus
```{r}
physeq_eb <- subset_samples(merged_physeq_phylum, Species == "Eb")
physeq_eb <- prune_taxa(taxa_sums(physeq_eb)>0, physeq_eb)
```

```{r}
ancom_rand_output_struc_eb = ancombc2(data = physeq_eb, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_eb$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("metagenomics"=2)

nrow(structural_zero_groups)
```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in metagenomics***
```{r}
structural_zero_groups %>% 
  filter(metagenomics==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Enriched phyla***
```{r}
ancom_rand_output_struc_eb$res 
```

#### Pipistrellus
```{r}
physeq_pk <- subset_samples(merged_physeq_phylum, Species == "Pk")
physeq_pk <- prune_taxa(taxa_sums(physeq_pk)>0, physeq_pk)
```

```{r}
ancom_rand_output_struc_pk = ancombc2(data = physeq_pk, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_pk$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("metagenomics"=2)

nrow(structural_zero_groups)
```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in metagenomics***
```{r}
structural_zero_groups %>% 
  filter(metagenomics==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Enriched phyla***
```{r}
ancom_rand_output_struc_pk$res %>% 
  filter(q_methodMetagenomics<0.05) %>%
  tt()
```

#### Hypsugo
```{r}
physeq_ha <- subset_samples(merged_physeq_phylum, Species == "Ha")
physeq_ha <- prune_taxa(taxa_sums(physeq_ha)>0, physeq_ha)
```

```{r}
ancom_rand_output_struc_ha = ancombc2(data = physeq_ha, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "method",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "method", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

***Total structural zeros***
```{r}
structural_zeros <- ancom_rand_output_struc_ha$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("amplicon"=1) %>% 
  rename("metagenomics"=2)

nrow(structural_zero_groups)
```

***Structural zeros in amplicon***
```{r}
structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(amplicon==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Structural zeros in metagenomics***
```{r}
structural_zero_groups %>% 
  filter(metagenomics==TRUE) %>% 
  rownames_to_column(., "phylum") %>% 
  left_join(., data.frame(physeq_eb@tax_table) %>% rownames_to_column(., "phylum"), by="phylum") %>% 
  select(phylum) %>% 
  pull()
```

***Enriched phyla***
```{r}
ancom_rand_output_struc_ha$res %>% 
  filter(q_methodMetagenomics<0.05) %>%
  tt()
```