# Functional differences

## Metagenomics

```{r load_data_func_met,comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("resources/metagenomics/data_filtered.Rdata")
```

```{r gift_analysis_meta, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),]

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

#Get community-weighed average GIFTs per sample
genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 
# genome_counts_row <- rownames_to_column(genome_counts_row, "genome")
GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)

unique_funct_db <- GIFT_db[c(3, 4, 5)] %>%
  distinct(Code_function, .keep_all = TRUE)
```

### Functional capacity of the MAGs

```{r GIFTs_elements_meta, comment="", message=FALSE, warning=FALSE, fig.dim = c(20, 20)}
GIFTs_elements %>%
  reshape2::melt() %>%
  dplyr::rename(Genome = Var1, Code_element = Var2, GIFT = value) %>%
  inner_join(GIFT_db,by="Code_element") %>%
  ggplot(., aes(x=Code_element, y=Genome, fill=GIFT, group=Code_function))+
  geom_tile()+
  scale_y_discrete(guide = guide_axis(check.overlap = TRUE))+
  scale_x_discrete(guide = guide_axis(check.overlap = TRUE))+
  scale_fill_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  facet_grid(. ~ Code_function, scales = "free", space = "free")+
  theme_grey(base_size=8)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),strip.text.x = element_text(angle = 90))
```

### MCI
```{r mci_meta, warning=FALSE, comments="", message=FALSE}
GIFTs_functions_community %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function)) %>% 
  group_by(Species, Domain) %>%
  summarise(mean = mean(value), sd = sd(value)) %>% 
  mutate(MCI = str_c(round(mean, 3), "±", round(sd, 3))) %>% 
  select(-mean, -sd) %>% 
  pivot_wider(names_from = Domain,
    values_from = MCI)
```

```{r gitfs_meta, message=FALSE, warning=FALSE}
unique_funct_db <- GIFT_db[c(3, 4, 5)] %>%
  distinct(Code_function, .keep_all = TRUE)

MCI <- GIFTs_functions_community %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function))

MCI_sub <- MCI %>% filter(Domain=="Biosynthesis")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Degradation")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Structure")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

```

***Functional differences considering the three species***
```{r commun_wilcox_elem_meta, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c(1,2)], by=join_by("sample"=="sample"))
```

***p adjusted***
```{r}
element_gift %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```

***p value***
```{r}
element_gift %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  filter(p_value < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```
```{r}
# element_gift <- GIFTs_elements_community %>% 
#   as.data.frame() %>% 
#   rownames_to_column(., "sample") %>% 
#   left_join(., sample_metadata[c(1,2)], by = join_by("sample" == "sample"))

# uniqueGIFT_db <- unique(GIFT_db[c(2,4,5,6)]) %>% 
#   unite("Function", Function:Element, sep = "_", remove = FALSE)

anova_results <- element_gift %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(anova_p = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]  ) %>%
#  mutate(p_adjust_ANOVA = p.adjust(anova_p, method = "BH")) %>%
  filter(anova_p < 0.05)

element_gift %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))

uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements <- element_gift %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))
```
```{r}
sig_Pk_Eb <- significant_elements %>% filter(contrast=="Pk-Eb")

filtered_elements_sig_Pk_Eb <- element_gift %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Eb$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Pk")) 


filtered_elements_sig_Pk_Eb %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}
sig_Pk_Ha <- significant_elements %>% filter(contrast=="Pk-Ha")

filtered_elements_sig_Pk_Ha <- element_gift %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Ha$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Ha"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Ha)%>% 
  mutate(group_color = ifelse(Difference <0, "Ha","Pk")) 

filtered_elements_sig_Pk_Ha %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#6b7398", "#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}
sig_Ha_Eb <- significant_elements %>% filter(contrast=="Ha-Eb")

filtered_elements_sig_Ha_Eb <- element_gift %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Ha_Eb$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Ha", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Ha-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Ha")) 

filtered_elements_sig_Ha_Eb %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#e5bd5b", "#6b7398")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```

## Metabarcoding Standard

```{r load_data_func_sta,comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("resources/amplicon/data_standard_filt.Rdata")
```

### Prepare for PicrusT

```{r prepare_picrust,comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
write_tsv(genome_counts_filt,"resources/amplicon/picrust/counts.tsv")
asvs <- genome_metadata$genome
fasta_all <- readDNAStringSet("resources/amplicon/ASVs_seqtab_overlap10_nochim0.fa", format="fasta")
fasta_filtered <- fasta_all[asvs]
writeXStringSet(fasta_filtered,
                filepath = "resources/amplicon/picrust/sequences.fa",
                format   = "fasta")
```

### Run Picrust

```{sh run_picrust,comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
conda activate picrust2
picrust2_pipeline.py -s resources/amplicon/picrust/sequences.fa \
      -i resources/amplicon/picrust/counts.tsv \
      -o resources/amplicon/picrust/results \
      -p 4 \
      --verbose
```

### Analyse functions

```{r load_functions,comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
amplicon_ko <- read_tsv("resources/amplicon/picrust/results/combined_KO_predicted.tsv")
amplicon_ec <- read_tsv("resources/amplicon/picrust/results/combined_EC_predicted.tsv")

tree_bacteria <- read.tree("resources/amplicon/picrust/results/bac_reduced.tre")
```

```{r distillr_functions,comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
amplicon_ko_long <- amplicon_ko  %>% 
  pivot_longer(!sequence,names_to="kegg",values_to="value") %>% 
  filter(value>0) %>% 
  mutate(kegg = stringr::str_remove(kegg, "^ko:"))

amplicon_ec_long <- amplicon_ec  %>% 
  pivot_longer(!sequence,names_to="ec",values_to="value") %>% 
  filter(value>0) %>% 
  mutate(ec = str_c("[EC:",ec,"]"))

# Distil KO and EC separately (for later merging)
amplicon_ko_distil <- distill(amplicon_ko_long,GIFT_db,genomecol=1,annotcol=2, verbosity=T)
amplicon_ec_distil <- distill(amplicon_ec_long,GIFT_db,genomecol=1,annotcol=2, verbosity=T)

# Save distilled tables
amplicon_ko_distil %>% 
  as.data.frame() %>% 
  rownames_to_column("genome") %>% 
  write_tsv("resources/amplicon/picrust/results/gifts_ko.tsv")

amplicon_ec_distil %>% 
  as.data.frame() %>% 
  rownames_to_column("genome") %>% 
  write_tsv("resources/amplicon/picrust/results/gifts_ec.tsv")
```

```{r distillr_functions_merge, comment="", echo=FALSE, message=FALSE, warning=FALSE}
# Merge distilled tables
amplicon_ko_distil <- read_tsv("resources/amplicon/picrust/results/gifts_ko.tsv") %>% column_to_rownames(var="genome")
amplicon_ec_distil <- read_tsv("resources/amplicon/picrust/results/gifts_ec.tsv") %>% column_to_rownames(var="genome")

amplicon_distil <- pmax(amplicon_ko_distil, amplicon_ec_distil)
amplicon_elements <- to.elements(amplicon_distil,GIFT_db)

#Aggregate element-level GIFTs into the function level
amplicon_functions <- to.functions(amplicon_elements,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 

amplicon_elem_community_standard <- to.community(amplicon_elements,genome_counts_row,GIFT_db)
amplicon_func_community_standard <- to.community(amplicon_functions,genome_counts_row,GIFT_db)
```

### MCI
```{r mci_sta, warning=FALSE, comments="", message=FALSE}
amplicon_func_community_standard %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function)) %>% 
  group_by(Species, Domain) %>%
  summarise(mean = mean(value), sd = sd(value)) %>% 
  mutate(MCI = str_c(round(mean, 3), "±", round(sd, 3))) %>% 
  select(-mean, -sd) %>% 
  pivot_wider(names_from = Domain,
    values_from = MCI)
```

```{r gitfs_sta, message=FALSE, warning=FALSE}


MCI <- amplicon_func_community_standard %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function))

MCI_sub <- MCI %>% filter(Domain=="Biosynthesis")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Degradation")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Structure")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

```

```{r distillr_functions_community_standard, comment="", echo=FALSE, message=FALSE, warning=FALSE}
amplicon_func_community_standard %>%
    as.data.frame() %>%
    rownames_to_column(var="sample")%>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata, by = join_by(sample == sample)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_grid(functionid ~ Species, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=8),
              strip.text.y = element_text(angle = 0)
              ) +
        labs(y="Traits",x="Samples",fill="GIFT")
```

***Functional differences considering the three species***
```{r commun_wilcox_elem_sta, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift_MS <- amplicon_elem_community_standard %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c("sample","Species")], by=join_by("sample"=="sample"))
```

***p adjusted***
```{r}
element_gift_MS %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```

***p value***
```{r}
element_gift_MS %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  filter(p_value < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```
```{r}
# element_gift <- GIFTs_elements_community %>% 
#   as.data.frame() %>% 
#   rownames_to_column(., "sample") %>% 
#   left_join(., sample_metadata[c(1,2)], by = join_by("sample" == "sample"))

# uniqueGIFT_db <- unique(GIFT_db[c(2,4,5,6)]) %>% 
#   unite("Function", Function:Element, sep = "_", remove = FALSE)

anova_results_MS <- element_gift_MS %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(anova_p = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]  ) %>%
#  mutate(p_adjust_ANOVA = p.adjust(anova_p, method = "BH")) %>%
  filter(anova_p < 0.05)

element_gift_MS %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results_MS$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))

uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements_MS <- element_gift_MS %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results_MS$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))
```

```{r}
sig_Pk_Eb_MS <- significant_elements_MS %>% filter(contrast=="Pk-Eb")

filtered_elements_sig_Pk_Eb_MS <- element_gift_MS %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Eb_MS$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Pk")) 


filtered_elements_sig_Pk_Eb_MS %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#e5bd5b", "#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}
sig_Pk_Ha_MS <- significant_elements_MS %>% filter(contrast=="Pk-Ha")

filtered_elements_sig_Pk_Ha_MS <- element_gift_MS %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Ha_MS$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Ha"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Ha)%>% 
  mutate(group_color = ifelse(Difference <0, "Ha","Pk")) 

filtered_elements_sig_Pk_Ha_MS %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#6b7398", "#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}
sig_Ha_Eb_MS <- significant_elements_MS %>% filter(contrast=="Ha-Eb")

filtered_elements_sig_Ha_Eb_MS <- element_gift_MS %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Ha_Eb_MS$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Ha", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Ha-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Ha")) 

filtered_elements_sig_Ha_Eb_MS %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#e5bd5b", "#6b7398")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```

### Metabarcoding selected

```{r distillr_functions_community_sel,comment="", echo=FALSE, message=FALSE, warning=FALSE}
load("resources/amplicon/selected_filtering.Rdata")

genome_counts_row <- genome_counts_filt_amplicon %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 

amplicon_elements_filtered <- amplicon_elements[rownames(genome_counts_row),]

#Aggregate element-level GIFTs into the function level
amplicon_functions <- to.functions(amplicon_elements_filtered,GIFT_db)

amplicon_elem_community_selected <- to.community(amplicon_elements_filtered,genome_counts_row,GIFT_db)
amplicon_func_community_selected <- to.community(amplicon_functions,genome_counts_row,GIFT_db)
```

```{r}
amplicon_elem_community_selected %>%
    as.data.frame() %>%
    rownames_to_column(var="sample")%>%
    pivot_longer(!sample,names_to="trait",values_to="gift") %>%
    left_join(sample_metadata_amplicon, by = join_by(sample == sample)) %>%
    mutate(functionid = substr(trait, 1, 3)) %>%
    mutate(trait = case_when(
      trait %in% GIFT_db$Code_element ~ GIFT_db$Element[match(trait, GIFT_db$Code_element)],
      TRUE ~ trait
    )) %>%
    mutate(functionid = case_when(
      functionid %in% GIFT_db$Code_function ~ GIFT_db$Function[match(functionid, GIFT_db$Code_function)],
      TRUE ~ functionid
    )) %>%
    mutate(trait=factor(trait,levels=unique(GIFT_db$Element))) %>%
    mutate(functionid=factor(functionid,levels=unique(GIFT_db$Function))) %>%
    ggplot(aes(x=sample,y=trait,fill=gift)) +
        geom_tile(colour="white", linewidth=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "#ddf1da")))+
        facet_grid(functionid ~ Species, scales="free",space="free") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              axis.text.y = element_text(size=8),
              strip.text.y = element_text(angle = 0)
              ) +
        labs(y="Traits",x="Samples",fill="GIFT")
```

### MCI
```{r mci_sel, warning=FALSE, comments="", message=FALSE}
amplicon_func_community_selected %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function)) %>% 
  group_by(Species, Domain) %>%
  summarise(mean = mean(value), sd = sd(value)) %>% 
  mutate(MCI = str_c(round(mean, 3), "±", round(sd, 3))) %>% 
  select(-mean, -sd) %>% 
  pivot_wider(names_from = Domain,
    values_from = MCI)
```

```{r gitfs_sel, message=FALSE, warning=FALSE}
unique_funct_db <- GIFT_db[c(3, 4, 5)] %>%
  distinct(Code_function, .keep_all = TRUE)

MCI <- amplicon_func_community_selected %>%
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>%
  pivot_longer(-sample, names_to = "trait", values_to = "value") %>%
  left_join(sample_metadata%>% select(sample, Species), by = join_by(sample == sample)) %>%
  left_join(unique_funct_db[c(1,2)], by = join_by(trait == Code_function))

MCI_sub <- MCI %>% filter(Domain=="Biosynthesis")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Degradation")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

MCI_sub <- MCI %>% filter(Domain=="Structure")
shapiro.test(MCI_sub$value)
kruskal.test(value ~ Species, data = MCI_sub)

```

***Functional differences considering the three species***
```{r commun_wilcox_elem_sel, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift_MSel <- amplicon_elem_community_selected %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata[c("sample","Species")], by=join_by("sample"=="sample"))
```

***p adjusted***
```{r}
element_gift_MSel %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```

***p value***
```{r}
element_gift_MSel %>%
  pivot_longer(-c(sample,Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]) %>%
  filter(p_value < 0.05) %>%
  left_join(.,uniqueGIFT_db[c(1,3)],by = join_by(trait == Code_element))
```
```{r}
# element_gift <- GIFTs_elements_community %>% 
#   as.data.frame() %>% 
#   rownames_to_column(., "sample") %>% 
#   left_join(., sample_metadata[c(1,2)], by = join_by("sample" == "sample"))

# uniqueGIFT_db <- unique(GIFT_db[c(2,4,5,6)]) %>% 
#   unite("Function", Function:Element, sep = "_", remove = FALSE)

anova_results_MSel <- element_gift_MSel %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(anova_p = summary(aov(value ~ Species))[[1]][["Pr(>F)"]][1]  ) %>%
#  mutate(p_adjust_ANOVA = p.adjust(anova_p, method = "BH")) %>%
  filter(anova_p < 0.05)

element_gift_MSel %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results_MSel$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))

uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_elements_MSel <- element_gift_MSel %>%
  pivot_longer(-c(sample, Species), names_to = "trait", values_to = "value") %>%
  filter(trait %in% anova_results_MSel$trait) %>% 
  group_by(trait) %>%
  summarise(
    tukey_results = list(
      as.data.frame(TukeyHSD(aov(value ~ Species))$Species) |> 
      tibble::rownames_to_column("contrast")
    )
  ) %>%
  unnest(tukey_results) %>% 
 filter(`p adj` < 0.05) %>% 
         left_join(., uniqueGIFT_db[c(1,3)], by = join_by(trait == Code_element))
```
```{r}
sig_Pk_Eb_MSel <- significant_elements_MSel %>% filter(contrast=="Pk-Eb")

filtered_elements_sig_Pk_Eb_MSel <- element_gift_MSel %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Eb_MSel$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Pk")) 


filtered_elements_sig_Pk_Eb_MSel %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#e5bd5b", "#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}
sig_Pk_Ha_MSel <- significant_elements_MSel %>% filter(contrast=="Pk-Ha")

filtered_elements_sig_Pk_Ha_MSel <- element_gift_MSel %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Pk_Ha_MSel$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Pk", "Ha"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Pk-Ha)%>% 
  mutate(group_color = ifelse(Difference <0, "Ha","Pk")) 

filtered_elements_sig_Pk_Ha_MSel %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#6b7398", "#76b183")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```
```{r}


sig_Ha_Eb_MSel <- significant_elements_MSel %>% filter(contrast=="Ha-Eb")

filtered_elements_sig_Ha_Eb_MSel <- element_gift_MSel %>%
  column_to_rownames(., "sample") %>% 
  select(sig_Ha_Eb_MSel$trait)%>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select("sample","Species"), by = join_by(sample == sample)) %>% 
  filter(Species %in% c("Ha", "Eb"))%>%
  select(-sample) %>%
  group_by(Species) %>%
  summarise(across(everything(), mean)) %>%
  t() %>%
  row_to_names(row_number = 1) %>%
  as.data.frame() %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column(., "Elements") %>%
  left_join(.,uniqueGIFT_db[c(1,3,4)],by = join_by(Elements == Code_element)) %>% 
  arrange(Function) %>% 
  mutate(Difference=Ha-Eb)%>% 
  mutate(group_color = ifelse(Difference <0, "Eb","Ha")) 

filtered_elements_sig_Ha_Eb_MSel %>%
  ggplot(aes(x=forcats::fct_reorder(Function,Difference), y=Difference, fill=group_color)) + 
  geom_col() +
#  geom_point(size=4) + 
  scale_fill_manual(values=c("#e5bd5b", "#6b7398")) + 
  geom_hline(yintercept=0) + 
  coord_flip()+
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        legend.title = element_blank(),
        panel.background = element_blank(),
          panel.grid.major = element_line(size = 0.15, linetype = 'solid',
                                colour = "grey"))+
  xlab("Function") + 
  ylab("Mean difference")
```